### 2017-09-28 Manuscript/replyToReferees.tex
Initial commit

\documentclass[letter,11pt,draft]{article}
\usepackage{color}
%\usepackage{hyphenat}
%\usepackage{soul}
%\linespread{1.0}
%\textwidth = 6.5 in
%\textheight = 8.5 in
%\oddsidemargin = 0.0 in
%\evensidemargin = 0.0 in
%\topmargin = 0.2 in
%\headheight = 0.2 in
%\headsep = 0.1 in
%\parskip = 0.15 in
%\parindent = 0.0in

\newcommand{\comment}[2]{{\textbf{Comment}~(#1):\it ``#2''}}
\newcommand{\co}[2]{{\textbf{Comment}~(#1):\it ``#2''}}
\newcommand{\reply}{\textbf{Reply}~}
\newcommand{\action}{\textbf{Action}~}
%\newcommand{\re}{\textbf{R}~}
\newcommand{\ac}{\textbf{A}~}
%\newcommand{\section}[1]{{\Large {\bfseries #1}}}
\newcommand{\re}[1]{\emph{\textcolor{blue}{#1}}}

\renewcommand{\b}[1]{\textcolor{blue}{#1}}

\begin{document}

%\begin{letter}{Cytometry A Editorial Office}

%\signature{Amir Erez, Robert Vogel, Andrew Mugler, Andrew Belmonte, and Gr\'egoire Altan-Bonnet}
%\signature{Amir Erez}

%\opening{Dear Editor,}

Dear Editor,

\bigskip

We thank the Referees for making many detailed and useful suggestions; we sincerely appreciate such a thorough review process.
Referee 1 found our example "illustrative and intuitive" for which "the proposed method clearly works well", whereas Referee 2 considered our work "worth bringing to the attention of the readership of Cytometry A" and that "the illustration seems particularly enlightening".

\bigskip
We have addressed all the Referee comments, point by point, below. A Referee comment appear in \re{blue italics}, followed by our reply in black. In addition, we have made the data and MATLAB code used for this manuscript freely available in REF HERE. We have also provided the MIFlowCyt details as supplements to this submission. We hope you will find our new submission much improved.

%\closing{We look forward to your reply,}
\bigskip
\bigskip
\bigskip
We look forward to your reply,

\bigskip
Kind Regards,

\bigskip
\bigskip
\bigskip
Amir Erez

\newpage
\section*{Reply to Associate Editor}

\re{
Associate Editor Comments:
\\
In addition to reviewers with respect to meeting scientific standards for reporting, it seems this work currently does not meet the MIFlowCyt requirements (not provided, as required by Cytometry A), especially as noted  with respect to data availability. Also missing are details of available implementations of their approach, per  Cytometry A software and data sharing requirements.}

\newpage

\section*{Reply to Referee 1}

We thank the Referee for considering our example \re{"illustrative and intuitive"} for which \re{"the proposed method clearly works well"}, and that the real-world example \re{"seems to do its job"}. We also thank the Referee for the multiple constructive suggestions. We detail each change following the Referee's comments below:

\smallskip

\re{1. It takes a reader while before it is clear what Q(I) and P(log I) mean in the introduction. They are formally introduced in the introduction, while it would be more coherent to do so in the Materials and Methods section. In addition, I would rewrite the introduction in a general non-formal way in order to make the problem statement clear to the reader.}

\smallskip
We appreciate the Referee's suggestion and moved the formal relation between Q(I) and P(log I) to the Methods section. We have also rewritten the opening of the introduction to more gently introduce the reader to the problem. We trust that the introduction now has a more general and non-formal flavor as suggested by the Referee.

\re{2. While more advanced data-transformations are stated in the introduction (e.g. Logicle transformation), it might be a good idea to also make at least a comment about them in the discussion of the work and how the method would relate to these transformations, as they are more and more the standard when analyzing flow cytometry data.}

\smallskip
We have added a comment about the more advanced transformations in the discussion. Please note that when analyzing the experimental data, in Fig.~3A, we show the data as presented by FlowJo using such a transformation. As is apparent in Fig.~3A, this does not solve the mismatch in the number of peaks.

\re{3. The 'Methods' section needs considerable restructuring. Preferably the toy (or simulated) example is pushed to the first part of the results section, and the methods section contains the derivation and structure of proposed methods, a brief description of Hartigan's dip test and a description of the used datasets.}

\smallskip
We have made particular effort to separate out the theoretical (methods) from the experimental (results). We find that in manuscripts that combine both, it is best to emphasize the difference, so that there is no doubt which is which, even at a glance. Whereas the Referee's suggestion has merit, it is also a question of style and suffers from trade-offs such as a less clear separation between experimental and theoretical. As such, we describe the used datasets where they are used, in the "Results - experiments" section. Since the two examples are different, we do so in series, so as to maintain focus on the relevant dataset. Further, in accordance with MIFlowCyt, the experimental details are summarized in the MIFlowCyt supplements. We hope the Referee understands and accepts our choice of presentation.

\re{4. There is no description of the data that is used in this study whatsoever. It is not clear whether the data from ref. [1] were reused or acquired in the same way as described in ref. [1]. Both datasets cannot be accessed. Details about the data are not present, which they should be. Preferably, they should also be made available, please consider Flowrepository to freely store your flow cytometry data (especially for ref. [1], as this data can only accessed upon request from correponding author, although they are published in an open-access journal).}

\smallskip
We have addressed this, providing description of the data in accordance with MIFlowCyt. We have made the data and scripts which create the figures publicly available. Please find the details in the Reply to Associate Editor.

\re{5. The link to phase transitions needs more explanation or to be removed (line 53-54).}

\smallskip
Removed.

\re{6. The second method, how biological information can be used to properly estimate the mismatch between logarithmic and linear binning is in need of a proper presentation, as the idea seems an interesting one, but is not clearly laid out.}

\smallskip
To clarify this method, we have added a paragraph, which begins with "We briefly sketch what follows". The purpose of this section is how, by measuring additional, related biological information, one can ameliorate the mismatch between the two representations, resulting in more reliable gates.

\re{7. A recent publication in Cytometry A by Johnsson et al. (2017) show that dip and bandwidth tests perform statistically consistent for unimodality, so it would be preferable to make a comment about that and to incorporate some of the literature in this manuscript. In addition, they propose that for the dip test you should have at least 10,000 cells while results in the manuscript suggest to have at least 100,000 cells. They analyzed the FlowCAP1-dataset, which might be a good addition to this work, as it is well known in the cytometry community.}
\\ 
\re{See: What is a “Unimodal” Cell Population? Using Statistical Tests as Criteria for Unimodality in Automated Gating and Quality Control. Johnsson et al. (2017), Cytometry part A.}

\smallskip
We thank the Referee for pointing out this interesting and recent manuscript which complements well the main thrust of our manuscript. As per Referee 2's suggestions, we have trimmed the detailed discussion of dip statistics (for which there are many, as Referee 2 points out), and focused on the mismatch instead. Accordingly, a deeper analysis of Johnsson et al. falls outside the focus of our work. Indeed, as Referee 2 points out, drawing conclusions about whether 10,000 or 100,000 cells are sufficient for a certain statistical test, depends on many factors and is not necessary for the message in our article. We cite Johnsson et al. in the manuscript in the appropriate place, together with related work.

\re{
8. Some parts of the manuscript, such as “the intuitive argument” on page 8, should be placed in the discussion.}

\smallskip
Moved. We thank the Referee and agree that it is better placed in the discussion.

\re{
9. Some labels are missing in figures. Please add those in figure 2 and 6. Also, a number of captions do not address figures in full, see for example the caption under figure 1, in which the left column is not addressed.}

\smallskip
We thank the Referee for pointing these out. We have done all changes as requested

\smallskip
\begin{itemize}
\item In Fig. 1: Caption  modified.
\item In Fig.2: $y\longrightarrow$ label added to the horizontal axis; the vertical axis shows 3 different functions, color-coded in the legend.
\item In Fig.5: (formerly Fig. 6) we added the appropriate labels.
\end{itemize}

\re{
10. Please have a look at the manuscript organization of Cytometry part A: http://onlinelibrary.wiley.com/journal/10.1002/(ISSN)1552-4930/homepage/ForAuthors.html.
}

\smallskip
We have made the appropriate changes in manuscript organization.

\smallskip
\re{
Minor comments:
\\
11. There are quite a number of spelling mistakes and grammatical errors in the manuscript.}

\smallskip
We are afraid that neither we nor the spell-checker in our software have found any obvious spelling mistakes. We have done our best to use correct grammar. We encourage the Referee to specifically point at mistakes they would like addressed.

\re{
12. Please adhere to the journal's format of abbreviations, such as flow cytometry (FCM).}

\smallskip
Fixed

\re{
13. Sometimes abbreviations are not defined such as 'RHS' and 'LHS' on page 5 L22 (presumably right hand side and left hand side).}

\smallskip
Fixed

\re{
14. Throughout document: please write 'unimodal' and 'bimodal'.
}

\smallskip
Changed.

\re{
15. The lines in figure 5 seem to be a bit 'clumsily' added; this figure would improve in quality if these lines were properly displayed.
}

\smallskip
The Referee refers to Fig.~4, (formerly Fig.~5). We're afraid we did not quite understand which lines have been 'clumsily' added, and what precisely does it mean. We gathered the Referee meant the plots of the marginal distributions $P(I_{ERK1})$ and $P(I_{ppERK})$ drawn in black and red (respectively) to the right of and on top of their joint heatmap. We have modified them, and agree with the Referee that the result is more aesthetic.

\re{
16. Please remove the last sentence of the discussion, as this is conjecture.
}

\smallskip
Removed.

\re{
17. The bibliography uses an inconsistent style. Please see the reference format of the journal.
}

\smallskip
Fixed.

\newpage

\section*{Reply to Referee 2}

We thank the Referee for considering our work \re{"worth bringing to the attention of the readership of Cytometry A"} and that \re{"the illustration seems particularly enlightening"}. We also thank the Referee for the multiple constructive suggestions and agree with the prediction that this more terse manuscript conveys the message more readily. We detail each change following the Referee's comments below:

\smallskip
\re{The notion and terminology employed in the manuscript is inconsistent with standard usage in probability and statistics, which is unfortunate as it is likely to limit/inhibit readers. I appreciate that changing this might be time consuming, though I personally think it would be worthwhile.
}
\re{What is labeled as Q(I) and called a distribution in the present paper is instead a probability density function (pdf). Let us say there is a single continuous random variable, I, the intensity of a fluorescence then the pdf would usually labeled $f_I(x)$, with associated cumulative distribution function (cdf) being $F_I(x) = int_{-\infty}^x f_I(y) dy$. What is called P(log I) would then the pdf of the random variable log(I), which would normally be written as $f_{log(I)}(x)$.}

\smallskip
We thank the Referee for the suggestion regarding precise notation of the pdfs in the manuscript, and that the Referee acknowledges that such a change of notation is a time consuming and personal choice. Certainly, for some of the readership of Cytometry A the suggested change of notation $Q(I)\rightarrow f_I(x)$ and $P(\log I)\rightarrow f_{\log I}(x)$ would improve readability. We fear, however, that a possibly comparable proportion of the readers might mistake $f_I$ and $f_{\log I}$ for having the same \emph{functional form}, just of a different argument, which is manifestly not the case in our example. Indeed, that is precisely the reason we chose $P$ and $Q$ to make this distinction explicit to the reader with a less formal mathematical background. More so, $P$, instead of $Q$, was granted to $f_{\log I}$ on merit that the reader is most likely acquainted with a logarithmic representation of the data, and is more habituated to a pdf denoted $P$. This choice of $P$ for the pdf of $\log I$ is consistent with Cytometry A's list of accepted abbreviations which chooses $P$ for probability.

\smallskip
We hope the Referee understands and accepts the thought process which led to our choice of notation.

\re{
That the relationship between the two is
\begin{equation}
   f_I(x) = f_{log(I)}(log(x))/x   
        \label{eq:changevar}         
\end{equation}
is due to the "change of variable" formula, which can be found in most undergraduate probability texts. }

\smallskip
We agree with this statement and indeed we have cited such a book by Nitis Mukhopadhyay in ref 25. The purpose of including this well-known relation is to introduce the reader less versed in probability theory with an essential component for the understanding of the manuscript.

\re{Immediately after Eq.~\ref{eq:changevar}  (i.e.  current eq. 1), is presented, I think one could pose the question: Q1: when is $f_I(x)$ uni-modal while $e^x f_I(e^x)=f_{log(I)}(x)$ is multi-modal and vice versa? That is the essential element of what follows, and the example provided makes the matter clear.}

\smallskip
We have added the question to the text as the Referee suggested.

\re{
Later in the paper where data is used to create empirical estimates of $f_I$ and $f_{log(I)}$, exactly how they are made should be stated by way of formulae and more standard notation employed. E.g. $\hat{f}^{(n)}_I$ where n is the sample size. Note that Eq.~\ref{eq:changevar} will not hold for most forms of estimation, which is why the distinction needs to be made.
}

\smallskip
We have added explicit details of how empirical densities are estimated in the relevant parts of the manuscript.

\re{
2) While some of the the argument in the paragraph starting "The difference..." is reasonable, it is not in its entirety and makes matters sound more frightening than they are. For many cell population enrichment purposes, there are gates (see all of hematopoiesis) that are based on logarithmic scales that are demonstrably segregating populations as established by cell sorting and downstream experiments.  Consequently, this broad brush-stroke dismissal of the meaning of log-scale separation seems at odds with established results.
}

\smallskip
We have changed the paragraph to address the Referee's concerns.

\re{
3) The methods section appears to mix up two things: a) when is Q1 (above) answered in the affirmative; b) when can you reliably assess this when the pdfs in question are not directly related as above, but are empirical estimates? The mixture of the two seems unnecessarily confused. In the present section, you are looking at the modeled pdf where there is no uncertainty so why would one need to confirm the computation with a (known to be unreliable) statistical test on simulated data?
\\
5) What is the relevance of the section on Hartigan's dip test? Tests for bimodality are notoriously challenging / conservative and all are established to be of limited power. There are plenty of references to that effect, so that is not a primary finding of the present paper. Presumably the purpose here is to point that well known challenge to this readership, which I think would be better stated explicitly, particularly as there is more than one dip test out there.}

\smallskip
We thank the Referee for pointing out this lack of focus in the methods section. The Referee is correct in understanding that some of our purpose here is to show that determining the number of peaks of experimental data is a difficult task. A second purpose is to refer to recent literature which employs Hartigan's test for a similar experimental situation. This is the relevance and particular choice of Hartigan as an example. Taking into account the Referee's comment, we have shortened the discussion of Hartigan's method in the main text, moving its bulk to the supplementary section for reference. We hope that in this way we clarify the focus of the methods section as per the Referee's suggestion.

\re{
4) In that section, the authors assume that $log(I)$ is a Gaussian mixture model. Tedious, unenlightening computations are then displayed much to this reader's displeasure, where only the outcome is necessary to be relayed. This is made more egregious due to the use of S (normally reserved for a survival function, i.e.  one minus the cdf, in all stats papers) and F (normally reserved for a cdf).  Essentially much of pages 4 and 5 could be deleted on the assumption that a sufficiently interested reader probably knows how to take derivatives and/or work mathematica, and the chase can be cut to.
}

\smallskip
We apologize for causing the Referee much displeasure. We accept the Referee's comment that the derivations do not need to be spelled out in full in the main body of the manuscript. Accordingly, we have moved some to the \emph{Supporting information} section, in our attempt to accommodate Cytometry A's readership which is of heterogeneous mathematical background. We are pleased to accommodate the Referee's suggestions that the letters $S$ and $F$ be left alone, and instead have changed $F\rightarrow A$ and $S_{1,3}\rightarrow B_{1,3}$, which we hope the Referee will find less objectionable.

\re{5) appears together with (3) above.}

\smallskip
\re{
6) The labels of Fig 3 are not self-consistent with the rest of the paper. What does events in $Q(I)$ mean given heretofore $Q(I)$ has been the probability density of the fluorescence at I? One presumes at this stage you are considering estimates of $f_I$. That being the case, a description of the empirical estimator would seem appropriate. In particular as one might note that (*) will no longer hold for these estimated pdfs (unless the pdf of one of I or log(I) is estimated and the transform used to estimate the other, which would be non-standard).
}

\smallskip
We have clarified the labels of Fig.~3 and have added a description of how the data are drawn from $Q(I)$.

\re{7) The notation used for a joint pdf, $P_2(...)$, is again not standard, and one would expect $f_{(log(I),log(J)}(x,y)$. Moreover, given the emphasis in the paper on the impact of change of variables on univariate random variables, it's surprising that there's no discussion of the multivariate equivalent.}

\smallskip
We hope the Referee understands that the same logic that guided our choice of $P,Q$ has led us to choose $P_2$, namely: (i) it is consistent with $P(\log I)$ in that $P$ is used to describe the pdf of the logarithm of intensity; (ii) $P_2$, instead of $P$, is used to make explicit the fact that these functions are different, not just of a different argument; (iii) $P$ is consistent with this journal's accepted abbreviations which specify $P$ for probability.

\smallskip
We have added a short comment about the multivariate change of variables, which lies outside the focus of the work. In our analysis, we focus on the marginal distributions which are univariate. We fear a deeper discussion of this will distract the readers from our main message.

\re{
8) From that section on, what was used in the present notation as a value a pdf is evaluated at, e.q. $Q(I)$, becomes a measured value.  Moreover the conditioning etc. is all written in a non-formal, non-standard way. Such abuse of notation is alleviated with use of standard notation, which would be advised.}

\smallskip
We refer the Referee to the book by Koller and Friedman (ref. 32 in the main text). Prof. Koller's book is widely acclaimed in the computational biology field, a highly cited standard reference (according to Google Scholar, 4438 citations since its publication in year 2009). The notation in the conditioning part of the manuscript matches that in Prof. Koller's book. The field of computational biology combines several disciplines with different notation conventions. We hope the Referee can accommodate this situation and accept our notation which is standard as is apparent in Prof. Koller's book.

\re{
A) One can't help but wonder if Q1 could be addressed in generality outside a parametric example.
\\
B) I expect that the authors will not wish to pay this too much heed, and I would not blame them for doing so, but mathematically speaking (rather than practically so) Q1 is not actually a sensible one. To be mathematically precise one would need to pose its equivalent for cdfs. The reason is that pdfs are only defined up to sets of Lebesgue measure 0 and so any unimodal pdf can be replaced with an arbitrarily multimodal one while still giving rise to the same cdf (e.g. spike a normal pdf with the value 100 at all rationales, and you will still have a normal cdf). In other words, a random variable does not have a single pdf and it is actually the case that one is defining equivalence classes through the cdf and picking a representative.
}

\smallskip
We thank the Referee for both these comments. We also wonder (i) whether the question can be addressed in generality; (ii) is there/how to find a class of representations that always preserve the number of peaks while maintaining the advantages of the logarithmic transform. These are, however, mathematical questions rather than immediately practical ones; this manuscript attempts to communicate with the cytometry community in a practical, readable manner. Such depth of mathematical analysis required to address the above is outside the scope of this journal. Once the review process is finished, we encourage the Referee to contact us if they so wish, to discuss these questions in more mathematical terms.

\re{C) Cytometry readings are made for many purposes. While the vast majority that I am aware of are on logarithmic scales, and indeed it is befuddling that they look beautifully log-normal, some are linear scales. For example forward and side scatter (FSC and SSC).}

\smallskip
We have added a comment regarding this to the discussion. Naturally, if no logarithmic transform is employed, then the possibility for mismatch disappears.

\re{
D) The sentence
"... similarly to the way Landau theory defines the critical point in second order phase transitions [28]"
seems to be an irrelevant remark for the present journal.}

\smallskip
Removed.

%\end{letter}

\end{document}

### 2017-09-28 Manuscript/LogSpaceDistributionsCytometryA.tex
Initial commit

%% Run LaTeX on this file several times to get Table of Contents,
%% cross-references, and citations.

\documentclass[11pt,a4paper,final]{article}
\usepackage{setspace}
\doublespacing
%\usepackage[top=0.85in,left=2.75in,footskip=0.75in]{geometry}

%\usepackage{Wiley-AuthoringTemplate}
%\usepackage[sectionbib,authoryear]{natbib}% for name-date citation comment the below line
%\usepackage[numbers]{natbib}% for numbered citation comment the above line
\usepackage[numbers,round]{natbib} % Round parenthesis

\raggedright
\setlength{\parindent}{0.5cm}
%\textwidth 5.25in 
\textheight 8.75in
\usepackage[aboveskip=1pt,labelfont=bf,labelsep=period,justification=raggedright,singlelinecheck=off]{caption}
\usepackage{graphicx}
\renewcommand{\figurename}{Fig}

\begin{document}
\def\be{\begin{equation}}
\def\ee{\end{equation}}
\def\bea{\begin{eqnarray}}
\def\eea{\end{eqnarray}}
\def\nl{\\ \noindent}
\def\nn{\\ \nonumber}
\def\etal{\emph{et al.} }
\newcommand{\gauss}[3]{e^{-\frac{(#3 - #2)^2}{2\sigma_{#1}^2}}}

\begin{flushleft}
{\Large
\textbf\newline{Modeling of cytometry data in logarithmic space: when is a bimodal distribution not bimodal?} % Please use "sentence case" for title and headings (capitalize only the first word in a title (or heading), the first word in a subtitle (or subheading), and any proper nouns).
}
\newline
% Insert author names, affiliations and corresponding author email (do not include titles, positions, or degrees).
\\
Amir Erez\textsuperscript{1*},
Robert Vogel\textsuperscript{2},
Andrew Mugler\textsuperscript{3},
Andrew Belmonte\textsuperscript{1,4},
Gr\'egoire Altan-Bonnet\textsuperscript{1}

\bigskip
\textbf{1} Immunodynamics Group, Cancer and Inflammation Program, National Cancer Institute, National Institutes of Health, Bethesda, Maryland 20814, USA
\\
\textbf{2} IBM T. J. Watson Research Center, Yorktown Heights, New York 10598, USA
\\
\textbf{3} Department of Physics and Astronomy, Purdue University, West Lafayette, Indiana 47907, USA
\\
\textbf{4} Department of Mathematics, Pennsylvania State University, University Park, Pennsylvania 16802, USA
\bigskip

% Use the asterisk to denote corresponding authorship and provide email address in note below.
* erezam@bgu.ac.il
\end{flushleft}

\section*{Acknowledgments}
This work was supported by Human Frontier Science Program grant LT000123/2014 (Amir Erez) and by the Intramural Research Program of the NCI, NIH.

\newpage

\begin{abstract}
Recent efforts in systems immunology lead researchers to build quantitative models of cell activation and differentiation. One goal is to account for the distributions of proteins from single-cell measurements by flow cytometry or mass cytometry as a readout of biological regulation. In that context, large cell-to-cell variability is often observed in biological quantities. We show here that these readouts, viewed in logarithmic scale may result in two easily-distinguishable modes, while the underlying distribution (in linear scale) is unimodal. We introduce a simple mathematical test to highlight this mismatch. We then dissect the flow of influence of cell-to-cell variability using a graphical model and its effect on measurement noise. Finally we show how acquiring additional biological information can be used to reduce uncertainty introduced by cell-to-cell variability, helping to clarify whether the data is uni- or bimodal. 
This communication has cautionary implications for manual and automatic gating strategies, as well as clustering and modeling of single-cell measurements. 
\end{abstract}

keywords: bimodal, unimodal, peak, logarithm, gating, FCM, CyTOF

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\newpage
\section*{Introduction}

Flow cytometry (FCM) data typically stretches across several orders of magnitude, with fluorescence intensity $I$ readily spanning values between $10^2$ and $10^5$. Although individual cells (events) are recorded as fluorescence intensity $I$, typically what is displayed to the investigator is a binned or smoothed histogram which represents an empirical probability density function (pdf) of the data. Due to the wide span of the data, binning directly the intensity $I$ is noisy since there is no single scale one can apply. Instead, when binning FCM data to create a histogram representing a pdf, it is natural to let bin sizes increase as a geometric progression, namely, to evenly bin the logarithm of the fluorescence intensity. As a result, instead of the probability $Q(I)$ (pdf) of fluorescence intensity $I$, one usually analyzes the probability (pdf) of $\log I$, which we denote $P(\log I)$. Indeed, $P(\log I)$ has many advantages: easy display of many orders of magnitude in $I$, easy to model as a two-component log-normal mixture model (as in \cite{Vogel2016}), and easy to intuitively understand the effect of changing the voltage gain on the flow-cytometer detector photo-multiplier. While such data presentation has been widely adopted in the field of cytometry out of these practical reasons, a rigorous assessment of this log-transformation reveals unwarranted features.

\smallskip
Simply plotting $Q(I)$ vs. $I$ is impractical as most of the data inevitably appears crowded against the $I=0$ axis. Thus, it is common practice to plot $P(\log I)$ or variants thereof which deal with small and negative $I$ values introduced by fluorescence compensation ({\it e.g.} "Logicle" \cite{Parks2006}, "VLog" \cite{Bagwell2016} and other transformations \cite{Finak2010}). Displaying faithfully FCM data is not easy, as the logarithmic scale and fluorescence compensation introduce problems that are easy to miss \cite{Herzenberg2006} leading to uncertainty in the number of distinct populations present in the data. Previously, attention has been given to the possibility of effects produced by logarithmic binning \cite{Novo2008}, contrasting the difference between plotting logarithmic histograms $P(\log I)$ vs. $\log I$ as opposed to rescaling the x-axis by plotting $Q(I)$ vs. $\log I$. However, an additional, potentially confusing situation seems to have been overlooked: the possible appearance of a second mode in $P(\log I)$, rendering $P(\log I)$ bimodal, while for the same data only one mode exists in $Q(I)$. This is the focus of this work.
\smallskip

When considering biological measurements, $I$ is proportional to the \emph{actual} copy number of RNA or proteins. When theoretical considerations are applied to biological systems (such as biochemical dynamics \cite{Friedman2006,Prill2015,Pal2015,Ridden2015,Mojtahedi2016,Erez2017}, mass-action chemical equilibria, cell-cycle measurements \cite{Brown2010} and Hill dose-response curves \cite{Prinz2010}), it is the copy number itself that is under consideration. Despite that, the logarithm of copy number is an appealing quantity because of its approximately Gaussian statistics, yielding insight into details easily lost if the data were to be analyzed only in linear scale.  This leads to a mismatch, where for instance models posed in linear space and data plotted in logarithmic space seem unable to be reconciled without invoking additional effects such as stochastic gene expression noise \cite{Friedman2006} and cell-to-cell variability \cite{Feinerman2008,Pelkmans2012,Cotari2013}. Even so, typically one must resort to approximations to analyze noise propagation linearly \cite{Prill2015}.
\smallskip

The difference between the convenient consideration of the logarithm of abundances and the theoretically-accurate analysis of the linear copy number renders the question of whether $Q(I)$ has one or two modes (3 extrema) relevant in the following ways: (i) the existence of 1 or 3 extrema is often used to infer the fixed points of a dynamic stochastic biochemical network \cite{Friedman2006,Pal2015,Vogel2016} and other \emph{in silico} methods \cite{Das2009} (ii) extrema are used to define cell-types in automatic (density based) gating and clustering algorithms \cite{Cron2013,ONeill2013,Anchang2014,Finak2014,Saeys2016,Chen2016}; (iii) The existence of a clearly bimodal distribution is used for manual gating ({\it e.g.} discerning between activated and un-activated cells) in a way that may appear more robust and compelling than it might truly be. Yet, immunologists have relied on FCM for over forty years to identify new cell populations, often based on manual gating of FCM data. In particular, in cell population enrichment experiments, often such markers are chosen that the (positive) signal is sufficiently well separated from the (negative) background for the two (logarithmic, linear) representations to have no mismatch in their number of peaks. At some level, these experiments succeed in gating the FCM data because typically only these high signal-to-noise markers are used, based on prior knowledge. However, not all FCM experiments enjoy such favorable conditions. Moreover, as the field advances, and more and more markers are acquired, some of these will inevitably have lower signal to noise resolution. The purpose of this manuscript is not to claim that \emph{all} logarithmic gates are problematic. Instead, we caution here of a situation where peaks that appear sufficiently separated in logarithmic space may yet mislead. We propose a simple test to warn if for a given sample, this situations arises.

\smallskip
The rest of this paper is composed of two parts. In the first part, we point out and analyze the situation where a mismatch between the two representations can happen: we formally state the problem using theoretical modeling of FCM data as a mixture of two log-normal distributions (colloquially the "negative" and "positive" modes), explicitly show situations where two modes appear in $P(\log I)$ while only one mode exists in $Q(I)$, and demonstrate this confounding effect on experimental data. In the second part, we analyze the role of cell-to-cell variability in experimental data and show how by measuring a suitable extra dimension one can factor out some of this variability, thus reducing the broadness of the modes sufficiently so as to reduce the mismatch between the two representations. Thus we provide a prescription to design experiments and analyze them so as to resolve the unimodal {\it vs.} bimodal discrepancy.

\newpage

\section*{Methods - theory}
Given the pdf $P(\log I)$, one can formally derive $Q(I)$ as \cite{Mukhopadhyay2000},
\bea 
\label{eq:P2Q}
Q(I) &=& P(\log I) \left\vert \frac{d}{dI} \log I \right\vert \nn
 &=& \frac{1}{I}P(\log I) = e^{-y} P(y)\,,
\eea
with $\log I\equiv y$. This relation serves us both in this theoretical section and in the experimental \emph{Results} section, where we estimate the empirical distribution for $Q(I)$ using this relation. We now address the question: when is $Q(I)$ unimodal while $P({\log I})$ is multimodal ?
\smallskip

Cytometry data is often amenable to modeling as a log-normal mixture ({\it e.g.} \cite{Vogel2016}). To demonstrate the log/linear mismatch we consider a mixture of two populations, characterized by the distribution of intensity. We define $P(\log I)$ as follows:
\be
\label{eq:LogNormalMix}
P(\log I) = \frac{(1-\alpha)}{\sqrt{2\pi\sigma_1^2}}e^{-\frac{(\log I - y_1)^2}{2\sigma_1^2}} + \frac{\alpha}{\sqrt{2\pi\sigma_2^2}}e^{-\frac{(\log I - y_2)^2}{2\sigma_2^2}},
\ee
with $y_{1,2}=\log I_{1,2}$ which are the loci of the centers of the left and right Gaussians in log-space, respectively, and $\sigma_{1,2}$ the log-space standard deviations. We then define $Q(I)=\frac{1}{I}P(\log I)$ as in Eq.~\ref{eq:P2Q}. 
\smallskip

In Fig.~\ref{fig:PlotsLinearLog}, to illustrate with typical measurement values, we set $I_1=100$ and $I_2=1000$ (arbitrary units) and $\alpha=0.5$, while varying $\sigma_1=\sigma_2$. This figure presents the three cases we wish to contrast: on the left column, both $P(\log I)$ and $Q(I)$ are bimodal; in the central column, $P(\log I)$ is bimodal whereas $Q(I)$ is unimodal; on the right column, both $P(\log I)$ and $Q(I)$ are unimodal, a situation which we examine in more detail in Eq.~\ref{eq:Plogy3Maxima}. Moreover, we see an example where even when both are bimodal (left column), the loci of the modes are different. 
\smallskip

\begin{figure}[!ht]
 \centering
  \includegraphics[clip,width=1\hsize]{Fig1.png}
  \caption{\textbf{Log-normal mixture showing the mismatch in the number of peaks.} \textbf{Top row, red:} $P(\log I)$ vs. $\log I$ normalized to max. \textbf{Middle row, magenta:} $Q(I)$ vs. $I$ normalized to max, and plotted on a narrower range. \textbf{Bottom row, blue:} $Q(I)$ vs. $\log I$ normalized to max, rescaling the x-axis as in Ref.~\cite{Novo2008}. In the central column where $\sigma_1=\sigma_2=0.8$, $P(\log I)$ shows explicit bimodality whereas $Q(I)$ is unimodal. \textbf{Right column:} in this case, the variance $\sigma^2$ is large enough (see Eq.~\ref{eq:Plogy3Maxima}) that $P(\log I)$ has only one mode even though it is modeled as a mixture. In all cases: $I_1=100,\,I_2=1000,\,\alpha=0.5$ varying $\sigma_{1,2}=\{0.4,0.8,1.2\}$ (left to right columns, respectively). Leftmost column shows axes. }
  \label{fig:PlotsLinearLog}
\end{figure}

Determining whether empirical data is multimodal is a difficult task (eg., \cite{Silverman1981}). Recently, this issue has been revisited in detail \cite{Johnsson2017}. In the context of modeling FCM data \cite{Das2009}, it was addressed by using Hartigan's dip test for unimodality \cite{Hartigan1985}. Later, as a benchmark to our method we report the results of Hartigan's test on the experimental data. More details on Harigan's test can be found in the \emph{Supporting information}.
\smallskip

We define $y=\log I$ and proceed to find the number of extrema for $P(y)$ and $Q(I)$. We do this by recasting Eq.~\ref{eq:LogNormalMix} and its derivatives into the functions $A(y),B_1(y),B_3(y)$. Full mathematical details can be found in the \emph{Supporting information}. We introduce the following functions,
\bea 
\label{eq:extrema}
A(y) &=& \left(\frac{\alpha}{1-\alpha}\right)\left(\frac{\sigma_1}{\sigma_2}\right) e^{\frac{(y-y_1)^2}{2 \sigma_1^2}-\frac{(y-y_2)^2}{2 \sigma_2^2}}  \nn
B_1(y) &=& \frac{\frac{y-y_1}{\sigma_1^2}+1}{\frac{y_2-y}{\sigma_2^2}-1} \nn
B_3(y) &=& \frac{(y-y_1)/\sigma_1^2}{(y_2-y)/\sigma_2^2}\,.
\eea
To find the extrema of $Q(I)$, one must find the solutions $y^*$ for $B_1(y^*)=A(y^*)$. Similarly, for the extrema of $P(\log I)$ one must solve $B_3(y^*)=A(y^*)$.

\smallskip
$A(y)$ is the ratio of the two Gaussians in Eq.~\ref{eq:LogNormalMix} and is therefore always non-negative; this implies that any extremum $y_*$ must satisfy $B_1(y_*)\ge 0$ and $B_3(y_*)\ge 0$. We note that the condition for extrema in $Q(I)$ requires that $\frac{d}{dy}P(y)=P(y)$ whereas, of course, extremizing $P(\log I)$ sets its derivative to zero, demonstrating the fact that the loci of the modes for $P(\log I)$ and $Q(I)$ are manifestly different. The region where the log-space distribution shows a second mode occurs when for $B_3(y^*)=A(y^*)$ admits three solutions whereas $B_1(y^*)=A(y^*)$ admits only one solution. Given that these equations are transcendental, a graphical way to asses the number of solutions is to plot $A(y),B_1(y),B_3(y)$ and count the number of times $B_1$ and $B_3$ intersect $A$.\smallskip

In Fig.~\ref{fig:ExtremaSolutions}, we present an example of this graphical method. The mismatch between the number of extrema of $P(\log I)$ and $Q(I)$ is apparent whenever (red curve) $B_3(y)$ intersects $A$ at 3 points, whereas (blue curve) $B_1(y)$ only intersects $A$ once. \smallskip

\begin{figure}[!ht]
 \centering
  \includegraphics[clip,width=1\hsize]{Fig2.png}
  \caption{\textbf{Graphical solution to count the number of extrema.} When the red (blue) curves intersect the dashed black line, $P(\log I)$ ($Q(I)$) are extremized. Dashed black: $A(y)$, blue: $B_1(y)$ and red: $B_3(y)$. The mismatch between the number of extrema of $P(\log I)$ and $Q(I)$ is apparent when the red curve intersects $A$ at 3 points, whereas the blue curve only intersects $A$ once. In both cases, the loci of the extrema are different for the two distributions. The plots along the diagonal (which correspond to the cases in Fig.~\ref{fig:PlotsLinearLog}) show the case $\sigma_1=\sigma_2$ which simplifies $A(y)$ to a straight line (the general case being a parabola) in these axes.}
  \label{fig:ExtremaSolutions}
\end{figure}

In the plots along the diagonal, we have $\sigma_1=\sigma_2$ (as in Fig.~\ref{fig:PlotsLinearLog})  which simplifies $A(y)$ since the quadratic (Gaussian) terms cancel, leaving only an exponential. This leads to a simple criterion to determine whether $P(\log I)$ itself admits one or two modes - previously in Fig.~\ref{fig:PlotsLinearLog}(right) we saw an example where $P(\log I)$ is unimodal despite being generated from a mixture. Graphically, we see that for $B_3=F$ to have 3 solutions, $\log B_3(y)$ has to have a slope less than $\log A(y)$ about the extremum $y_*$. In other words, $\left.\frac{d}{dy} \log B_3(y)\right\vert_{y_*} \le \left.\frac{d}{dy} \log A(y)\right\vert_{y_*}$, with equality as the threshold between 1 and 3 extrema. This leads to the following intuitive criterion,
\be 
\label{eq:Plogy3Maxima}
(y_* - y_1)(y_2 - y_*) \ge \sigma ^2 \,\,\Longrightarrow \,\, \mbox{3 extrema for } P(\log I)\,,
\ee
which states that for $P(\log I)$ to appear bimodal, it must have an extremum ($y_*$) such that the variance of the individual Gaussian components of $P(\log I)$ must be smaller than the distance between $y_*$ and the Gaussian centers. Substituting for $y_* \approx \log 316$, $y_1 = \log 100$ and $y_2=\log 1000$ and $\sigma^2 = 1.44$ we see that the criterion in Eq.~\ref{eq:Plogy3Maxima} is not satisfied and indeed in Fig.~\ref{fig:PlotsLinearLog}(right) and Fig.~\ref{fig:ExtremaSolutions}(bottom right) we see that $P(\log I)$ has only one mode. A similar condition can be derived for $Q(I)$, that is, $\left.\frac{d}{dy} \log B_1(y)\right\vert_{y_*} \le \left.\frac{d}{dy} \log A(y)\right\vert_{y_*}$, such that,
\be 
\label{eq:QI3Maxima}
(y_* - y_1+\sigma^2)(y_2 - y_* - \sigma^2) \ge \sigma ^2\,\,\Longrightarrow \,\,\mbox{3 extrema for } Q(I)\,,
\ee \smallskip
it is, however, hard to compare the two bounds analytically because the $y_*$ which extremizes $P(\log I)$ is different from the $y_*$ which extremizes $Q(I)$.

\newpage

\section*{Results - experiments}

We now apply our analysis to experimental data, namely the measured distribution of Extracellular Signal-regulated Kinase (ERK) phosphorylation (ppERK) signaling in CD8+ primary mouse T-cells responding to antigens and inhibited by the SRC inhibitor Dasatinib. These data were acquired in exactly the same manner as the experiments in Ref.~\cite{Vogel2016}, for brevity we refer the reader there for all experimental details. In Fig.~\ref{fig:Experiment}(A) we see how a commercially-available analysis software (FlowJo \cite{FlowJo}) plots the distribution of ppERK in such an experiment, which clearly shows a bimodal structure. Fig.~\ref{fig:Experiment}(B) plots those same data, formally $P(\log I) := \hat{f}_{\log I}^{(8751)}$, the probability density estimated from the data by taking the logarithm of the intensity of 8751 cells and binning it in bins of size 0.2. Our procedure for estimating $P(\log I)$ gives the two modes as in FlowJo (red dots) whereas $Q(I)$ (derived using Eq.~\ref{eq:P2Q}) has a single mode (blue dots). We fit $P(\log I)$ as a Gaussian mixture. This is followed in Fig.~\ref{fig:Experiment}(C) by the same extrema analysis as in Fig.~\ref{fig:ExtremaSolutions}, revealing that indeed $Q(I)$ has a single maximum.
\smallskip

\begin{figure}[!ht]
 \centering
  \includegraphics[width=1\linewidth]{Fig3.png}
  \caption{\textbf{Analysis of experimental data reveals the effect we describe in a real scenario. }\textbf{(A)} Histogram of ppERK as plotted by FlowJo \cite{FlowJo}; \textbf{(B)} Histograms for $P(\log I)$ (red,dots) and $Q(I)$ (blue,dots) vs. $\log I$ as estimated from the data binned logarithmically. Note that plotting $Q(I)$ vs. $\log I$ is somewhat unusual but allows both $P$ and $Q$ to be plotted on the same axis. The red line shows the result of fitting $P(\log I)$ to a gaussian mixture model (Eq.~\ref{eq:LogNormalMix}), and the blue line is the estimate for $Q(I)$ from $P(\log I)$ according to Eq.~\ref{eq:P2Q}. The blue star indicates the location of the only maximum for $Q(I)$ obtained from Eq. \ref{eq:extremaQ}, despite the obvious two maxima in $P(\log I)$ (red). Hartigan's unimodality p-values for $\log I$ (red) and $I$ (blue) are taken directly from the data without binning, corroborating that $\log I$ is bimodal whereas $I$ is unimodal. \textbf{(C)} Graphic solution of the extrema conditions as in Fig.~\ref{fig:ExtremaSolutions} explicitly reveals the three solutions for $P(\log I_*)$ (red line intersects black dashed) as opposed to the single solution for $Q(I_*)$ (the blue line intersects the black dashed line up beyond the plotted area, solution also plotted as blue star in the middle plot), indicating that $Q(I)$ has only one mode.}
  \label{fig:Experiment}
\end{figure}
Given the prevalence and success of manual gating of FCM data in many situations, we wondered whether gating in logarithmic scale could be justified \emph{a posteriori}, based on biological knowledge. In the final part of our analysis, we aimed to include additional information such that the biological significance of the distributions in our single-cell measurements is better captured.

\smallskip
We briefly sketch what follows: by considering both ERK1 and ppERK levels, (the two being biologically related and experimentally correlated), we demonstrate that dividing one by the other ameliorates the mismatch between the logarithmic and linear representations. This is done by examining their \emph{joint} probability density function $P_2$; we suggest a probabilistic graphical model that describes the correlations in the system and their dependence on cell-to-cell variability; we suggest a way to normalize for cell-to-cell variability, such that the mismatch (resulting from increased uncertainty) between logarithmic and linear gates is resolved. We then generalize our example and provide a simple experimental and theoretical framework that allows less ambiguous determination of positive/negative gates.

\smallskip
In Fig.~\ref{fig:ppERKvsERK1heatmap}(A), we show the experimental data we will use in our proposed solution. Here we returned to our single-cell measurements of ERK phosphorylation in primary mouse T cells in Ref.~\cite{Feinerman2008}. We show a heat map of the \emph{joint} distribution of ppERK ($I_{ppERK}$) and total ERK1 ($I_{ERK1}$) expression in mouse CD8+ T-cells. We estimate their joint distribution by using a kernel density estimator \cite{botev2010}, which is used only for presentation purposes and for demonstrating their conditional independence (Fig.~\ref{S1_Fig}).
Notably, whereas the two modes of ppERK significantly overlap when plotted in the marginal distribution $P(\log I_{ppERK})$, ppERK expression correlates with total ERK1 levels in their joint distribution $P_2(\log I_{ppERK},\log I_{ERK1})$. The relation in Eq.~\ref{eq:P2Q} can be generalized to $P_2(\log I_{ppERK},\log I_{ERK1})$ (see, eg., \cite{Mukhopadhyay2000}), however, it will not be used in this manuscript.

\smallskip
To estimate the marginal densities $P(\log I_{ppERK})$ and $P(\log I_{ERK1})$ we bin the log-intensity of 24449 cells.  Each cell's state encodes another \emph{latent} variable, its activation status - which tells if the cell has been successfully activated by the stimulus. To deduce the activation status, it is common practice to use manual gating of the data by drawing a boundary between the active and inactive states. To account for the correlation between ERK1 and ppERK we consider two manual gating strategies: (i) perpendicular gating (dashed red) according to $P(\log I_{ppERK})$ with $I_{ppERK}>ppERK_*$ considered an activated cell, and (ii) diagonal gating according to the apparent correlation in $P_2$ (dashed grey). We set the diagonal gate with a slope of unity, meaning that we take the dividing line, reflecting  proportionality $ppERK\propto ERK1$, as a good way to partition the two states. We define "Inactive" to the left of the dashed line, and "Active" to the right of it.
\smallskip

To understand the structure of these data, it is important to characterize explicitly the dependency structure of our observables (ERK1, ppERK), the latent activation status, and the influence of external factors on these three. The existence of two peaks in ppERK which appear distinct from each other but correlated with ERK1 levels, guides us to use a Bayesian network to capture these features in the data as a graphical model. First - we test whether ERK1 and the cell's activation status are independent. In Fig.~\ref{S1_Fig} we see that whereas independence implies that $P(\log ERK1) = P(\log ERK1 \vert Activation)$, in fact there is a weak dependence between them regardless of whether we employ a vertical (red) or a diagonal (grey) gate (the diagonal gate showing a weaker dependence). The weak dependence between activation state and ERK1 levels is reasonable, if we account for cell-to-cell variability, since for a given stimulus some cells inevitably respond differently from the typical cell \cite{Spencer2009}. We summarize the causal structure for this system in Fig.~\ref{fig:ppERKvsERK1heatmap}(B) which depicts a probabilistic graphical model \cite{Koller2009} of the flow of influence from cell-to-cell variability and activation signal, to ERK1 levels and activation status, and finally to the distribution of ppERK. We depict the pair ERK1-ppERK in a template (rectangles), to suggest to the reader the existence of multiple other pairs. Importantly, in what follows we show how to better resolve the log-space peak; this recipe, together with the model in Fig.~\ref{fig:ppERKvsERK1heatmap}(B) can be used \emph{a priori} in automatic gating and clustering algorithms to prevent some of the mismatch between logarithmic and linear binning strategies. For stochastic modeling, such a structure presents an opportunity to analyze the structure and propagation of noise in the system \cite{Prill2015,Ching2017}.
\smallskip

We treat the broadness of ppERK modes as generated by cell-to-cell variability in total ERK1 content - a reasonable assumption since the noise in the phosphorylation of ERK is negligible in comparison \cite{Filippi2016}. We further neglect the indirect influence between activation status and ERK1 levels due to its weakness (checked in Fig.~\ref{S1_Fig}). Thus we approximate that the conditional independence between ERK1 and activation status (given that both are influenced by cell-to-cell variability) is true independence. This implies an approximately linear relation $I_{ppERK} \propto I_{ERK1}$ given activation status.
%This gives the following linear relation between ERK1 and ppERK,
%\begin{align}
%    \ce{ERK1 <=> MEK + ERK1 -> ppERK} \nn
%    \ce{ppERK -> ERK1}
%\end{align}
We define the normalized intensity $\tilde{I} = I_{ppERK}/I_{ERK1}$ as the ratio of ppERK to ERK1 intensity, thereby eliminating the linear dependence of ppERK on ERK1 levels and reducing uncertainty due to cell-to-cell variability. The resulting $P(\log \tilde{I})$ may boast a sufficiently reduced noise in ppERK such that a clear bimodal signature appears regardless of logarithmic or linear binning of $\tilde{I}$. In Fig.~\ref{fig:NormppERKbyERK} we show such an example, where in Fig.~\ref{fig:NormppERKbyERK}(A,B), $P(\log I)$ and $Q(I)$ do not agree on the number of modes, whereas in Fig.~\ref{fig:NormppERKbyERK}(C,D) $\tilde{I} = I_{ppERK} / I_{ERK1}$ do agree. These data have order 25,000 events and so, similarly to Fig.~\ref{fig:CheckHartigan}, Hartigan's test may not identify the number of peaks correctly, as is indicated in the $p_u$ values.

Thus we demonstrate how by suitably accounting for cell-to-cell variability one can reduce the measured noise so as to circumvent the mismatch in the number of modes between the logarithmic and linear treatment. For testing bimodality, whereas our method relies on fitting a Gaussian mixture, existing statistical tests, eg., Hartigan's test, require no fitting yet may lack statistical power when applied to typical experimental situations. 
\smallskip

\begin{figure}
    \centering
    \includegraphics[width=1\linewidth]{Fig4.png}
   \caption{\textbf{Analysis of experimental data with two correlated measurements.} \textbf{(A)} The joint distribution $P_2(\log I_{ppERK},\log I_{ERK1})$ as a heat map with its marginals plotted on the top and on its right. The correlation between ppERK and ERK1 levels is clear in the data. Dashed red (grey) lines are proposed manual gates according to the marginal (joint) distributions $P(\log I_{ppERK})$ $\left(P_2(\log I_{ppERK},\log I_{ERK1})\right)$. \textbf{(B)} Bayesian network depicted as a graphical model to show the flow of influence on the measurement of ppERK. The pair ERK1 and ppERK are in a template to suggest that there exist other pairs of correlated observables that depend on activation status and cell-to-cell variability.}
   \label{fig:ppERKvsERK1heatmap}
\end{figure}

\smallskip
\begin{figure}
 \centering
   \includegraphics[width=1\linewidth]{Fig5.png}  
  \caption{\textbf{By dividing ppERK readings by ERK1, we can ameliorate the mismatch between the two representations.} \textbf{(A)} $P(\log I_{ppERK})$ and $Q(I_{ppERK})$ vs. $\log I_{ppERK}$ (dots: data, lines: Gaussian mixture fit) together with \textbf{(B)} their extrema analysis, showing that the second mode in log ppERK does not exist if the data is linearly binned. \textbf{(C)} The same treatment but for $\tilde{I} = I_{ppERK}/I_{ERK1}$, \textbf{(D)} shows that both $\tilde{I}$ and $\log \tilde{I}$ have two modes, thus normalizing ppERK levels by total ERK1 maintains the bimodal structure both in $P(\log \tilde{I})$ and in $Q(\tilde{I})$. }
  \label{fig:NormppERKbyERK}
\end{figure}

\smallskip

\newpage
\section*{Discussion}
We conclude our analysis by making a more intuitive argument. Above, we have demonstrated the theoretical and experimental existence of a situation where $P(\log I)$ has two modes when $Q(I)$ has only one. But how does this \emph{come about} ? Intuitively, by binning in logarithmic scale, we are effectively making the bin sizes grow as $I$ increases. A larger bin can only lead to a higher count in that bin and so we might stumble upon a regime where this creates a quasi-mode. This situation and the same logic applies also when using more advanced logarithmic-like transformations, (eg., Logicle \cite{Parks2006} and Vlog \cite{Bagwell2016} ), as is demonstrated in Fig.~\ref{fig:Experiment}A.

%Probing further, we established a simple criterion (Eq.~\ref{eq:Plogy3Maxima}) where by making the underlying Gaussian mixture composed of too-close-together Gaussians, even in $P(\log I)$ there exists only one peak.

\smallskip
The scale-dependent bimodality as demonstrated in Fig.~\ref{fig:Experiment} and Fig.~\ref{fig:NormppERKbyERK}(A,B) may be not uncommon. Specifically, one must take extra care when attempting to manually gate, automatically cluster or build dynamical models which rely on an apparent bimodal structure, as it might depend on whether the data was log-transformed or not. This becomes increasingly relevant as cytometry moves forward to higher dimensional measurements which become tractable only with automatic gating schemes. Instead, one might consider plotting $Q(I)$ on the log-log scale, a presentation which preserves the number of maxima, at the expense of the measure of the distribution. It is possible to ameliorate the mismatch between the two scales, as we demonstrate in Fig.~\ref{fig:NormppERKbyERK}(C,D), if one can simultaneously measure correlated observables (in our example, ppERK and ERK1). This allows to control for cell-to-cell variability, increasing the resolution of the data. Recently, this favorable scenario has become more attainable with the introduction of mass cytometry - where one can rely on a large number of channels without compromising the FCM panel. Based on the analysis carried out in this paper, we conjecture that such extra channels, chosen wisely, can provide automatic clustering/gating algorithms the right information needed to make more reliable clustering and population defining. This is a simple way to introduce knowledge of the biological structure of the data into otherwise objective clustering algorithms, without compromising their objectivity. We propose and test some features of a graphical model that captures the structure of such dependencies in a way potentially useful for those interested in automatic gating and clustering algorithms.

\smallskip 
Logarithmic transform of FCM intensity values is the usual practice, but for some observations, eg., the FSC and SSC channels, linear scale is typically used. In such cases, there is no danger of mismatch between presentation. When FCM data are logarithmically distributed, though we caution on the use of $P(\log I)$, we find it remarkable how well the distribution of biological quantities resembles a log-normal mixture. 
%This highlights the deep and still little understood connection between distributions observed in living things and their relation to the logarithm of abundance, a subject likely to puzzle researchers for years to come.

\section*{Conflicts of Interest}
The authors declare that there are no conflicts of interest.

%\bibliographystyle{unsrt}
\bibliographystyle{cytA}

\bibliography{LogSpaceLatexPaper}

\newpage
\section*{Supporting information}
%
%\paragraph*{S1 Fig}
%\label{S1_Fig}

\subsection*{Finding the extrema of a log-normal mixture}
It is possible to discern between one or three extrema of the distribution $P(\log I)$, corresponding to one or two modes (respectively) by counting the number of solutions for $\frac{d}{d\log I}P(\log I) = 0$. Similarly, $\frac{d}{dI}Q(I)=0$ can have either one or three solutions. This raises the possibility of there being three extrema (two modes) for $P(\log I)$ whereas only one mode in $Q(I)$. We explicitly evaluate the extrema of the mixture of log-normal distributions by solving,
%
\bea \nonumber
  \left.\frac{d P(y)}{dy} \right\vert_{y=y_*}&=&
  -\frac{(y_* - y_1)}{\sigma_1^2} \frac{(1-\alpha)}{\sqrt{2\pi}\sigma_1}  \gauss{1}{y_1}{y} -
  \frac{(y_* - y_2)}{\sigma_2^2} \frac{(\alpha)}{\sqrt{2\pi}\sigma_2} \gauss{2}{y_2}{y}\nonumber\\
  &=& 0\nonumber
\eea
%
which by algebraic rearrangement,
%
\be 
  \label{eq:extremaP}
  \underbrace{\frac{(y_* - y_1) / \sigma_1^2}{(y_2-y_*)/\sigma_2^2}}_{B_3(y_*)} =
    \underbrace{\frac{\sigma_1}{\sigma_2}
    \left(\frac{\alpha}{1-\alpha}\right)
    e^{\frac{(y_*-y1)^2}{2\sigma_1 ^2}}\gauss{2}{y_2}{y_*}}_{A(y_*)}\,,
\ee
%
provides a more transparent form.

Here we refer to the left-hand-side and right-hand-side as $B_3(y)$ and $A(y)$, respectively. In like, computing the extrema of the linear scale distribution amounts to $\left.\frac{d Q(I)}{dI} \right\vert_{\log I= y_*} = 0$, which by change of variable is equivalent to,
%
\be
\label{eq:QExtremaStrange}
P(y_*) = \left.\frac{d P(y)}{dy}\right\vert_{y = y_*}\,,
\ee
%
and by substitution,
%
\be
  \label{eq:extremaQ}
  \underbrace{\frac{\frac{y_*-y_1}{\sigma_1^2}+1}{\frac{y_2-y_*}{\sigma_2^2}-1}}_{B_1(y_*)} = A(y_*) 
\ee
%
We refer to the quantity on the left-hand-side, $B_1(y)$.

\subsection*{Hartigan's dip test}
Briefly, Hartigan's dip statistic measures the maximum difference between the empirical distribution and the unimodal distribution that minimizes that maximum difference. This is compared to the appropriate null distribution which is, in this case, the uniform distribution, to give $p_u$, a p-value for unimodality.

As a check for the predictive power of Hartigan's test with regards to experimental data, we apply it on a log-normal mixture comparing its predictive power as a function of the number of tests and number of events in each test \cite{NicPrice}. For this, we draw events from $Q(I)$ by using MATLAB's \emph{lognrnd} function for each of the two log-normal distributions. In Fig.~\ref{fig:CheckHartigan}, we test it on the situation in Fig.~\ref{fig:ExtremaSolutions}(top,middle) in which $Q(I)$ is weakly bimodal, meaning that its bimodality is nearly marginal ($\sigma_1=0.4$ and $\sigma_2=0.8$). The Hartigan probability of unimodality ($p_u$) is not sensitive to the number of bootstrap tests in a reasonable range but becomes strongly predictive of the (weak) bimodality only when there are more than $10^5$ events. Such an abundance of cells may not always be available in typical FCM data, especially for sub-populations which have been selected (gated) and may comprise only a small fraction of all the cells acquired. Fig.~\ref{fig:CheckHartigan} supports that in such weakly bimodal situations, Hartigan's p-value should be treated cautiously, a situation which we encounter in Fig.~\ref{fig:NormppERKbyERK}. \smallskip

\begin{figure}[!ht]
 \centering
 \includegraphics[clip,width=1\hsize]{Fig6.png}
  \caption{\textbf{Testing Hartigan's p-value $p_u$ for unimodality on a weakly bimodal log-normal mixture.} The heat-map shows $p_u$ as a function of the number of bootstrap tests and number of events in each test, for $\sigma_1=0.4$ and $\sigma_2=0.8$ as in Fig.~\ref{fig:ExtremaSolutions}(top,middle). With less than $\sim10^5$ events, Hartigan's test for this case may misidentify the number of peaks.}
  \label{fig:CheckHartigan}
\end{figure}

\subsection*{Relation between ERK1 level and activation status}
\begin{figure}[!htb]
 \centering
   \includegraphics[width=1\linewidth]{Fig7.png}  
  \caption{{\bf Test for weak dependence of ERK1 and activation status.} Whereas independence implies that $P(\log ERK1) = P(\log ERK1 \vert Activation)$, in fact there is a weak dependence between them regardless of whether we employ a vertical (red, defined by threshold value $ppERK_*$) or a diagonal (grey) gate (the diagonal gate showing a weaker dependence); this is observed directly by noting that the different distributions in Fig.~\ref{S1_Fig} do not lie on top of each other. To quantify this difference, the inset shows the mutual information between $P(\log ERK1)$ and $P(\log ERK1 | Activation)$, with the circle pointing out the particular concentration of stimulus (out of all concentrations used in Ref.~\cite{Feinerman2008}) we chose to plot in this example. The chosen concentration has the highest mutual information, {\it i.e.}, the lowest ability to discern between the active and inactive states.}
  %\caption{\textbf{By dividing ppERK readings by ERK1, we can ameliorate the mismatch between the two representations.} \textbf{(A)} $P(\log I)$ and $Q(I)$ vs. $\log I$ (dots: data, lines: gaussian mixture fit) together with \textbf{(B)} their extrema analysis, showing that the second mode in log ppERK does not exist if the data is linearly binned. \textbf{(C)} The same treatment but for $\tilde{I} = I_{ppERK}/I_{ERK1}$, \textbf{(D)} shows that both $\tilde{I}$ and $\log \tilde{I}$ have two modes, thus normalizing ppERK levels by total ERK1 maintains the bimodal structure both in $P(\log \tilde{I})$ and in $Q(\tilde{I})$. }
  \label{S1_Fig}
\end{figure}

\end{document}

### 2017-09-28 Data/FeinermanScience2008/ERK1[072407]/ERK1.m

@@ -1,0 +1,147 @@
+
+clear all
+close all
+
+'Check file number & column order & check ppERk gate'
+
+NumberOfFiles=8;
+NumberOfGates=30;
+A=dir;
+concentrations=logspace(-8,-8-NumberOfFiles+1,NumberOfFiles);
+
+step=0.1;
+x=-1:step:1;
+y=x;
+[XI,YI]=meshgrid(x,y);
+
+
+for i=1:NumberOfFiles
+    clear tempdata
+    tempdata=[]
+    for j=1:size(A,1)
+        if findstr(['_00',num2str(i)], A(j).name)>0
+        A(j).name
+        Atemp=load(A(j).name);
+        tempdata=[tempdata' Atemp']';
+        end
+    end
+   
+    
+   length=size(tempdata,1);     
+   tempdata2=tempdata(1+0*floor(length/3):3*floor(length/3),:);
+    
+   MeanData=mean(log10(tempdata2));
+   
+   data{i}(:,1)=log10(tempdata2(:,3))-MeanData(3);
+   data{i}(:,2)=log10(tempdata2(:,4))-MeanData(4);
+   data{i}(:,3)=tempdata2(:,6)>tempdata2(:,4)/1.5;
+    
+ 
+   SortedData{i}=sortrows(data{i},2);
+  
+   EventsPerGate=floor(size(data{i},1)/NumberOfGates);
+    
+    for j=1:NumberOfGates
+            ppERK(i,j)=100*mean(SortedData{i}((j-1)*EventsPerGate+1:j*EventsPerGate,3),1);
+            FoldChange(j)=mean(SortedData{i}((j-1)*EventsPerGate+1:j*EventsPerGate,2),1);
+    end        
+    
+    [XI,YI,ZI{i}]=griddata(data{i}(:,2),data{i}(:,1),data{i}(:,3),XI,YI);
+    
+    subplot(3,3,i)
+    image(x,y,ZI{i}*64)
+    axis xy
+    xlabel('ERK-1')
+    ylabel('CD8\alpha')
+    zlabel('ppERK (red=no response / blue = response)')
+    title(['[OVA] = 10^{',num2str(-7-i),'} Mol']);
+    
+    
+end
+
+concentrations2=logspace(-8,-8-NumberOfFiles+1,21);
+
+
+                   
+figure
+for j=1:NumberOfGates
+    [estimates, model,sse] = Hill_fit(concentrations,ppERK(:,j)') 
+       
+    Amplitude2(j)=estimates(1);
[…]

### 2017-09-28 Data/FeinermanScience2008/ERK1[072407]/Hill_fit.m

@@ -1,0 +1,26 @@
+function [estimates, model,sse] = Hill_fit(concentrations,Frequency_response)
+
+
+% Call fminsearch with a random starting point.;
+%start_point = [max(Frequency_response)-min(Frequency_response) concentrations(floor(length(concentrations)/2)) min(Frequency_response)];
+start_point = [100 concentrations(floor(length(concentrations)/2)) min(Frequency_response)];
+model = @Hill_curve;
+estimates = fminsearch(model, start_point);
+[sse, FittedCurve] = Hill_curve(estimates);
+
+% Hill_curve accepts curve parameters as inputs, and outputs sse,
+% the sum of squares error for A * exp(-lambda * xdata) - ydata, 
+% and the FittedCurve. FMINSEARCH only needs sse, but we want to 
+% plot the FittedCurve at the end.
+    function [sse, FittedCurve] = Hill_curve(params)
+        
+        
+        length (concentrations);
+        A = params(1);
+        c0 = params(2);
+        b=params(3);
+        FittedCurve = b+A .* (concentrations)./(concentrations+c0);
+        ErrorVector = FittedCurve - Frequency_response;
+        sse = sum(ErrorVector .^ 2);
+    end
+end

### 2017-09-28 Data/FeinermanScience2008/ERK1[072407]/Hill_fit2.m

@@ -1,0 +1,25 @@
+function [estimates, model,sse] = Hill_fit(concentrations,Frequency_response)
+
+
+% Call fminsearch with a random starting point.;
+start_point = [max(Frequency_response)-min(Frequency_response) concentrations(floor(length(concentrations)/2)) min(Frequency_response)];
+model = @Hill_curve;
+estimates = fminsearch(model, start_point);
+[sse, FittedCurve] = Hill_curve(estimates);
+
+% Hill_curve accepts curve parameters as inputs, and outputs sse,
+% the sum of squares error for A * exp(-lambda * xdata) - ydata, 
+% and the FittedCurve. FMINSEARCH only needs sse, but we want to 
+% plot the FittedCurve at the end.
+    function [sse, FittedCurve] = Hill_curve(params)
+        
+        
+        length (concentrations);
+        A = params(1);
+        c0 = params(2);
+        b=params(3);
+        FittedCurve = (concentrations)./(concentrations+c0);
+        ErrorVector = FittedCurve - Frequency_response;
+        sse = sum(ErrorVector .^ 2);
+    end
+end

### 2017-09-28 Scripts/CopyFigures.sh

@@ -1,0 +1,8 @@
+cp Plots/PlotsLinearLogTogether.png ../Manuscript/Fig1.png
+cp Plots/ExtremaSolutions.png ../Manuscript/Fig2.png
+cp Plots/ExperimentHistograms.png ../Manuscript/Fig3B.png
+cp Plots/ExperimentExtremaAnalysis.png ../Manuscript/Fig3C.png
+cp Plots/FeinermanHeatmap.png ../Manuscript/Fig4A.png
+cp Plots/FixMismatch.png ../Manuscript/Fig5.png
+cp Plots/CheckHartigan.png ../Manuscript/Fig6.png
+cp Plots/PERK1Comparison.png ../Manuscript/Fig7.png

### 2017-09-28 Scripts/LogspacePaperPlots.m

@@ -1,0 +1,484 @@
+% Makes plots for logspace paper
+
+%% Plot histograms in log space and linear space 
+
+sig1s=0.4:0.4:1.2;
+sig2s=0.4:0.4:1.2;
+
+p.alpha = 0.5;
+p.y1 = log(100);
+p.y2 = log(1000);
+
+P = @(y,p)( (1-p.alpha)/(p.sig1*sqrt(2*pi))*exp(-((y-p.y1).^2)/(2*p.sig1^2)) + p.alpha/(p.sig2*sqrt(2*pi))*exp(-((y-p.y2).^2)/(2*p.sig2^2)));
+Q = @(y,p)(exp(-y).*P(y,p));
+figure;
+for s1=1:length(sig1s)
+    for s2=1:length(sig2s)
+        yl = [10^-5, 10^5];
+        yrange = log([10^0,10^5]);
+        ys = linspace(yrange(1),yrange(2),1001);
+        p.sig1 = sig1s(s1);
+        p.sig2 = sig2s(s2);
+        
+        subplot(3,3,s2+(s1-1)*length(sig2s)); 
+        valQ = Q(ys,p);
+        valQ = valQ / max(valQ(~isinf(valQ)));
+        semilogx(exp(ys),valQ,'-b','LineWidth',2);
+        hold on
+        set(gca,'FontSize',14);   
+        valP = P(ys,p);
+        valP = valP / max(valP(~isinf(valP)));
+        semilogx(exp(ys),valP,'-r','LineWidth',2);
+        xlim(exp(yrange));
+        title(['$\sigma_1=' num2str(sig1s(s1)) ', \sigma_2=' num2str(sig2s(s2)) '$'],'Interpreter','Latex');
+        set(gca,'XTick',10.^(1:2:5));
+        set(gca,'YTick',[]);
+    end
+end
+
+figname = 'PlotsLinearLogTogether';
+set(gcf,'Name',figname);
+% print(gcf,'-dtiff',['Plots' filesep figname],'-r600');
+% print(gcf,'-depsc2',['Plots' filesep figname]);
+close(gcf);
+
+%% Plot histograms in log space and linear space separately
+
+sig1s=0.4:0.4:1.2;
+sig2s=0.4:0.4:1.2;
+
+p=struct;
+p.alpha = 0.5;
+p.y1 = log(100);
+p.y2 = log(1000);
+
[…]

### 2017-09-28 Scripts/ReadERKData_Amir.m

@@ -1,0 +1,534 @@
+% Reads ERK data and plots
+
+A=dir;
+nFiles = 8;
+concentrations=logspace(-8,-8-nFiles+1,nFiles);
+
+
+filenames = cell(nFiles,1);
+Data = cell(nFiles,1);
+for ff=1:nFiles
+    filenames{ff} = ['../Data/FeinermanScience2008/ERK1[072407]/ERK1(OVA)_Tube_00' num2str(ff) '.txt'];
+    try
+        tabData = readtable(filenames{ff},'Delimiter','\t','ReadVariableNames',false);
+    catch
+        error(['Failed: file ' filenames{ff}]);
+    end
+    
+    tabData.Properties.VariableNames{1} = 'FCSA';
+    tabData.Properties.VariableNames{2} = 'SSCA';
+    tabData.Properties.VariableNames{3} = 'CD8a';
+    tabData.Properties.VariableNames{4} = 'ERK1';            
+    tabData.Properties.VariableNames{5} = 'DAPI';
+    tabData.Properties.VariableNames{6} = 'ppERK';
+    Data{ff} = tabData;    
+end
+
+%% Plot histogram of ppERK
+log10bn = [0:0.1:6];
+figure;
+hold on
+cmap = colormap(parula(nFiles));
+for ff=1:nFiles
+   hst = hist(log10(Data{ff}.ppERK),log10bn);
+%    hst = hist(log10(Data{ff}.ERK1),log10bn);
+   hst = hst / trapz(log10bn,hst);
+   plot(log10bn,hst,'.-','Color',cmap(ff,:),'DisplayName',num2str(concentrations(ff)));
+end
+
+%% Kernel density estimate of joint ppERK/ERK1 distribution
+
+% for ff=1:nFiles
+
+xcrit = 120; % This is where to slice the histogram
+chosen_dose = 4;
+
+% for ff=1:length(Data)
+for ff=chosen_dose:chosen_dose
+    data = log10(Data{ff}{:,{'ppERK','ERK1'}});
+    [bandwidth,density,X,Y]=kde2d(data);
+    % plot the data and the density estimate
+    figure;
+    set(gcf,'Units','normalized');
+    set(gcf,'Color','w');
+    set(gcf,'InvertHardcopy','off');
+    set(gcf,'Name',['Concentration 10^{' num2str(log10(concentrations(ff))) '}']);
[…]

### 2017-09-28 Scripts/hartigansdiptestdemo.m

@@ -1,0 +1,23 @@
+
+%create some obviously unimodal and bimodal Gaussian distributions just to
+%see what dip statistic does
+% Nic Price 2006
+
+Cent1 = ones(1,9);
+Cent2 = 1:1:9;
+sig = 0.5;
+nboot = 500;
+
+tic
+figure;
+for a = 1:length(Cent1),
+    xpdf(a,:) = sort([Cent1(a)+randn(1,200) Cent2(a)+randn(1,200)]); %allocate 200 points in each
+    [dip(a), p(a)] = HartigansDipSignifTest(xpdf(a,:), nboot);
+    
+    subplot(3,3,a)
+    hist(xpdf(a,:),-2:0.25:12)    
+    title(['dip=',num2str(dip(a),3), ', p=',num2str(p(a),3)])
+    xlim([-2 12])
+end
+% FixAxis([-2 12]);
+toc

### 2017-09-29 Manuscript/replyToReferees.tex
First main commit

@@
 \usepackage{color}
+\usepackage{hyperref}
 %\usepackage{hyphenat}
@@
 A Referee comment appear in \re{blue italics}, followed by our reply in black.
-In addition, we have made the data and MATLAB code used for this manuscript freely available in REF HERE.
+In addition, we have made the data and MATLAB code used for this manuscript freely available (details below).
 We have also provided the MIFlowCyt details as supplements to this submission.
@@
 
+\smallskip
+Dear Associate Editor, we now make publicly available all data (including MIFlowCyt checklists) and analysis with the manuscript, organized as described below.
+We have added to the \emph{Supplemental information} the description below.
+
+\begin{itemize}
+\item All MATLAB scripts used to generate the plots and analyze the data in
+\\
+{\small \href{https://github.com/AmirErez/BimodalLogspaceCytA/tree/master/Scripts}{https://github.com/AmirErez/BimodalLogspaceCytA/tree/master/Scripts}}
+\item Al data used in the manuscript and accompanying MIFlowCyt checklists, which comes from two datasets:
+	\begin{enumerate}
+	\item 20140920-OT1-dynamics - used for Fig. 3 in the manuscript.
+	\item FeinermanScience2008 - previously published in Feinerman et al. Science Vol. 321, Issue 5892, pp. 1081-1084 (2008). 
+	\end{enumerate}
+These data with their respective MIFlowCyt checklists are freely available in
+
+\smallskip
+\noindent{\small
+\href{https://github.com/AmirErez/BimodalLogspaceCytA/tree/master/Data/20140920-OT1-dynamics}{https://github.com/AmirErez/BimodalLogspaceCytA/tree/master/Data}}
+
+\end{itemize}
+
+\smallskip
+The main manuscript github page can be found in
+
+\smallskip
+\href{https://github.com/AmirErez/BimodalLogspaceCytA}{https://github.com/AmirErez/BimodalLogspaceCytA}
 \newpage
@@
 We find that in manuscripts that combine both, it is best to emphasize the difference, so that there is no doubt which is which, even at a glance.
-Whereas the Referee's suggestion has merit, it is also a question of style and suffers from trade-offs such as a less clear separation between experimental and theoretical.
 As such, we describe the used datasets where they are used, in the "Results - experiments" section.
-Since the two examples are different, we do so in series, so as to maintain focus on the relevant dataset.
-Further, in accordance with MIFlowCyt, the experimental details are summarized in the MIFlowCyt supplements.
+Further, in accordance with MIFlowCyt, the experimental details are summarized in the MIFlowCyt supplements and the data made freely available.
+We have added a comment to this effect at the beginning of the \emph{Methods} section and details of the data availability in the \emph{Supporting information}.
 We hope the Referee understands and accepts our choice of presentation.
@@
 \smallskip
-We have addressed this, providing description of the data in accordance with MIFlowCyt.
-We have made the data and scripts which create the figures publicly available.
-Please find the details in the Reply to Associate Editor.
+We have addressed this, providing description of the data in accordance with MIFlowCyt. We have made the data and scripts which create the figures publicly available. Please find the details above.
 
+\smallskip
 \re{5. The link to phase transitions needs more explanation or to be removed (line 53-54).}

### 2017-09-29 Manuscript/LogSpaceDistributionsCytometryA.tex
First main commit

@@
 \usepackage{setspace}
+\usepackage{hyperref}
 \doublespacing
@@
 \section*{Methods - theory}
+All data and MATLAB scripts used for the experimental part of this work, together with accompanying MIFlowCyt checklists detailing the data analysis, are freely available online.
+For details we refer the reader to the \emph{Supporting information}.
+The rest of this \emph{Methods} section deals with the theoretical part of this manuscript.
+
+\smallskip
 Given the pdf $P(\log I)$, one can formally derive $Q(I)$ as \cite{Mukhopadhyay2000},
@@
 
+\subsection*{Data and analysis}
+All data and analysis used for this manuscript has been made publicly available as follows:
+
+\smallskip
+\begin{itemize}
+\item All MATLAB scripts used to generate the plots and analyze the data in,
+\\
+{\small \href{https://github.com/AmirErez/BimodalLogspaceCytA/tree/master/Scripts}{https://github.com/AmirErez/BimodalLogspaceCytA/tree/master/Scripts}}
+\item Al data used in the manuscript and accompanying MIFlowCyt checklists, which comes from two datasets:
+	\begin{enumerate}
+	\item 20140920-OT1-dynamics - used for Fig. 3 in the manuscript.
+	\item FeinermanScience2008 - previously published in Feinerman et al. Science Vol. 321, Issue 5892, pp. 1081-1084 (2008). 
+	\end{enumerate}
+These data with their respective MIFlowCyt checklists are freely available in,
+
+\smallskip
+\noindent{\small
+\href{https://github.com/AmirErez/BimodalLogspaceCytA/tree/master/Data/20140920-OT1-dynamics}{https://github.com/AmirErez/BimodalLogspaceCytA/tree/master/Data}}
+
+\end{itemize}
+
+\smallskip
+The main manuscript github page can be found in,
+
+\smallskip
+\href{https://github.com/AmirErez/BimodalLogspaceCytA}{https://github.com/AmirErez/BimodalLogspaceCytA}
+
 \subsection*{Finding the extrema of a log-normal mixture}

### 2017-10-04 Manuscript/replyToReferees.tex
Some changes

@@
 \smallskip
-We have addressed this, providing description of the data in accordance with MIFlowCyt. We have made the data and scripts which create the figures publicly available. Please find the details above.
+We have addressed this, providing description of the data in accordance with MIFlowCyt.
+We have made the data and scripts which create the figures publicly available.
+We provide details in the \emph{Reply to Associate Editor}.
+Briefly, all data, scripts, and MIFlowCyt details can be found in
+\href{https://github.com/AmirErez/BimodalLogspaceCytA}{https://github.com/AmirErez/BimodalLogspaceCytA}
 
@@
 \smallskip
-To clarify this method, we have added a paragraph, which begins with "We briefly sketch what follows".
-The purpose of this section is how, by measuring additional, related biological information, one can ameliorate the mismatch between the two representations, resulting in more reliable gates.
+To clarify this method, we have added a paragraph, which begins with "We briefly sketch what follows". (Last paragraph in page 12).
+The purpose of this section is how, by considering both ERK1 and ppERK levels, one can ameliorate the mismatch between the two representations, resulting in more reliable gates.
 
@@
 \smallskip
-We are afraid that neither we nor the spell-checker in our software have found any obvious spelling mistakes.
-We have done our best to use correct grammar.
-We encourage the Referee to specifically point at mistakes they would like addressed.
+We thank the editor for catching these mistakes, and hope that these errors did not obscure his or her first reading of the manuscript.
+We have taken great care in revising this version and are happy to work with the editor and / or representatives from Cytometry A to correct any remaining errors.
 
@@
 \smallskip
-We thank the Referee for the suggestion regarding precise notation of the pdfs in the manuscript, and that the Referee acknowledges that such a change of notation is a time consuming and personal choice.
+We thank the Referee for the suggestion regarding precise notation of the pdfs in the manuscript, and that the Referee acknowledges that such a change of notation is a time consuming.
 Certainly, for some of the readership of Cytometry A the suggested change of notation $Q(I)\rightarrow f_I(x)$ and $P(\log I)\rightarrow f_{\log I}(x)$ would improve readability.

### 2017-11-08 Manuscript/LogSpaceDistributionsCytometryA.bak.tex
After second round of review

--- 2017-09-29 Manuscript/LogSpaceDistributionsCytometryA.tex
@@
   \includegraphics[clip,width=1\hsize]{Fig2.png}
-  \caption{\textbf{Graphical solution to count the number of extrema.} When the red (blue) curves intersect the dashed black line, $P(\log I)$ ($Q(I)$) are extremized.
+  \caption{\textbf{Graphical solution to count the number of extrema.} We test our peak counting method, based on the same equations as in Fig.~\ref{fig:PlotsLinearLog}, but with additional settings.
+When the red (blue) curves intersect the dashed black line, $P(\log I)$ ($Q(I)$) are extremized.
 Dashed black: $A(y)$, blue: $B_1(y)$ and red: $B_3(y)$.

### 2017-11-08 RefReports/replyToReferees.tex
Moved files

--- 2017-10-04 Manuscript/replyToReferees.tex
@@
 {\small \href{https://github.com/AmirErez/BimodalLogspaceCytA/tree/master/Scripts}{https://github.com/AmirErez/BimodalLogspaceCytA/tree/master/Scripts}}
-\item Al data used in the manuscript and accompanying MIFlowCyt checklists, which comes from two datasets:
+\item All data used in the manuscript and accompanying MIFlowCyt checklists, which comes from two datasets:
 	\begin{enumerate}
@@
 We have also rewritten the opening of the introduction to more gently introduce the reader to the problem.
-We trust that the introduction now has a more general and non-formal flavor as suggested by the Referee.
 
@@
 We have made particular effort to separate out the theoretical (methods) from the experimental (results).
-We find that in manuscripts that combine both, it is best to emphasize the difference, so that there is no doubt which is which, even at a glance.
 As such, we describe the used datasets where they are used, in the "Results - experiments" section.
@@
 We have added a comment to this effect at the beginning of the \emph{Methods} section and details of the data availability in the \emph{Supporting information}.
-We hope the Referee understands and accepts our choice of presentation.
 
@@
 The Referee refers to Fig.~4, (formerly Fig.~5).
-We're afraid we did not quite understand which lines have been 'clumsily' added, and what precisely does it mean.
 We gathered the Referee meant the plots of the marginal distributions $P(I_{ERK1})$ and $P(I_{ppERK})$ drawn in black and red (respectively) to the right of and on top of their joint heatmap.
@@
 \smallskip
-We apologize for causing the Referee much displeasure.
 We accept the Referee's comment that the derivations do not need to be spelled out in full in the main body of the manuscript.
@@
 \smallskip
+We believe that we have used standard notation.
 We refer the Referee to the book by Koller and Friedman (ref.
@@
 The field of computational biology combines several disciplines with different notation conventions.
-We hope the Referee can accommodate this situation and accept our notation which is standard as is apparent in Prof.
-Koller's book.
 

### 2017-11-08 RefReports/coverletterCytA.tex
Moved files

\documentclass[11pt]{letter}
\usepackage{color}
\usepackage{hyphenat}
\usepackage{soul}
\linespread{1.0}
\textwidth = 6.5 in
\textheight = 8.5 in
\oddsidemargin = 0.0 in
\evensidemargin = 0.0 in
\topmargin = 0.0 in
\headheight = 0.0 in
\headsep = 0.0 in
\parskip = 0.15 in
\parindent = 0.0in

\newcommand{\comment}[2]{{\textbf{Comment}~(#1):\it ``#2''}}
\newcommand{\co}[2]{{\textbf{Comment}~(#1):\it ``#2''}}
\newcommand{\reply}{\textbf{Reply}~}
\newcommand{\action}{\textbf{Action}~}
\newcommand{\re}{\textbf{R}~}
\newcommand{\ac}{\textbf{A}~}

\renewcommand{\b}[1]{\textcolor{blue}{#1}}

\begin{document}

\begin{letter}{Cytometry A Editorial Office}

%\signature{Amir Erez, Robert Vogel, Andrew Mugler, Andrew Belmonte, and Gr\'egoire Altan-Bonnet}
\signature{Amir Erez}

\opening{Dear Editor,}

Please find enclosed our manuscript ``Modeling of cytometry data in logarithmic space: when is a bimodal distribution not bimodal?'' submitted for consideration in {\it Cytometry A}.

Flow- and mass-cytometry have become essential tools in immunology and beyond. These measurement devices give abundant, highly sensitive single-cell data yielding distributions of cellular markers. These distributions are typically analyzed in logarithmic scale on a bi-axial plot where one selects (``gates") phenotypic subsets, thus defining distinct populations. As the technology progressed, it is now possible to measure a large number of dimensions, making manual analysis more difficult, as a 15-marker panel involves over 100 pairwise plots. Accordingly, the field moves to rely on automatic clustering and dimensional-reduction algorithms, sometimes rendering the analysis steps opaque. At the same time, advances in computational biology have yielded interesting mathematical models that are applied on such data, without always checking how logarithmically transforming the data might affect the models.

Here, we point out and analyze a mathematical phenomenon whereby distributions may appear bi-modal when binned in logarithmic scale, while appearing uni-modal in linear scale. Importantly, linear scale is where both mathematical models and actual DNA transcription and translation occurs. Thus we are faced with an inconvenient situation where models in linear scale and analysis in logarithmic scale may rely on density peaks which exist only in the logarithmic representation. Density-based automatic gating and clustering algorithms may be similarly confused by this mismatch. Indeed, technical changes in the experiment (eg. swapping one fluorescent dye with another, brighter one) might yield qualitatively different results despite being biologically equivalent, thus potentially causing reproducibility issues.

Our study advances the analysis of flow-and mass-cytometry data in several specific ways:

\begin{enumerate}
\item We explicitly analyze the occurrence of mismatch in the number of density peaks between the logarithmic and linear representations, and derive a mathematical test that indicates these conditions. We apply this test on experimental data, showing that the situation we describe may not be uncommon.
\item We compare our test with an existing statistical test, Hartigan's dip test, which has been employed previously in computational biology studies to determine bi-modality of cytometry distributions. Our test is more reliable when ran on a small subset of cells, less than approximately $10^5$, a typical situation for sub-populations of flow data.
\item We explore the effect of cell-to-cell variability on creating conditions where such a mismatch is likely. We then propose a way to design experiments and analyze data so as to circumvent the mismatch between the two representations, thus increasing the robustness and reproducibility of conclusions drawn.
\end{enumerate}

We believe this manuscript is quite appropriate for {\it Cytometry A}, and will be particularly useful to its readership, and more broadly to the growing community concerned with quantitatively correct and robust analysis of cytometry data and experimental design. Our manuscript highlights, dissects and ameliorates a situation that is not uncommon in many experiments, although we do not address any specific biological question. Moreover, as the readership of {\it Cytometry A} is more mathematically inclined than the biological community in general, they are likely the precise audience to fully grasp the implications of the mismatch we present, as well as to be involved in the development of models, automatically gating procedures, or cluster high-dimensional data in ways susceptible to these effects.

%We request that the following researchers be excluded from the review process due to current or recent collaboration or conflict of interest:
%\begin{itemize}
%\item Pankaj Mehta
%\item Ilya Nemenman
%\item Aleksandra Walczak
%\end{itemize}

\closing{We look forward to your reply,}

\end{letter}

\end{document}

### 2017-11-17 Manuscript/LogSpaceDistributionsCytometryA.tex
Changed figures to tif
tif
Update

--- 2017-09-28 Manuscript/LogSpaceDistributionsCytometryA.tex
@@
 \\
-Amir Erez\textsuperscript{1*},
+Amir Erez\textsuperscript{1*,5},
 Robert Vogel\textsuperscript{2},
@@
 \textbf{4} Department of Mathematics, Pennsylvania State University, University Park, Pennsylvania 16802, USA
+\\
+\textbf{5} Present address: Department of Molecular Biology, Princeton University, Lewis Thomas Laboratory, Princeton, New Jersey 08544, USA
 \bigskip
@@
 % Use the asterisk to denote corresponding authorship and provide email address in note below.
-* erezam@bgu.ac.il
+* aerez@princeton.edu
 \end{flushleft}
@@
 We introduce a simple mathematical test to highlight this mismatch.
-We then dissect the flow of influence of cell-to-cell variability using a graphical model and its effect on measurement noise.
+We then dissect the flow of influence of cell-to-cell variability proposing a graphical model which motivates higher-dimensional analysis of the data.
 Finally we show how acquiring additional biological information can be used to reduce uncertainty introduced by cell-to-cell variability, helping to clarify whether the data is uni- or bimodal.
@@
 \section*{Methods - theory}
+We focus this \emph{Methods} section on the theoretical aspect of this work. Brief details regarding the experimental methods can be found at the end of this section.
+
+\smallskip
 Given the pdf $P(\log I)$, one can formally derive $Q(I)$ as \cite{Mukhopadhyay2000},
@@
 We define $y=\log I$ and proceed to find the number of extrema for $P(y)$ and $Q(I)$.
-We do this by recasting Eq.~\ref{eq:LogNormalMix} and its derivatives into the functions $A(y),B_1(y),B_3(y)$.
-Full mathematical details can be found in the \emph{Supporting information}.
-We introduce the following functions,
+We do this by simple differentiation, demanding that the derivatives of $P$ and $Q$ equal zero as a definition of extrema.
+This is achieved by recasting Eq.~\ref{eq:LogNormalMix} and its derivatives into the functions $A(y),B_1(y),B_3(y)$, defined as,
 \bea 
@@
 \eea
-To find the extrema of $Q(I)$, one must find the solutions $y^*$ for $B_1(y^*)=A(y^*)$. Similarly, for the extrema of $P(\log I)$ one must solve $B_3(y^*)=A(y^*)$.
+The outcome of this approach is that, to find the extrema of $Q(I)$, one must find the solutions $y_{*q}$ for $B_1(y_{*q})=A(y_{*q})$.
+Similarly, for the extrema of $P(\log I)$ one must solve $B_3(y_{*p})=A(y_{*p})$.
+Full mathematical details on how these are derived can be found in the \emph{Supporting information}.
 
 \smallskip
-$A(y)$ is the ratio of the two Gaussians in Eq.~\ref{eq:LogNormalMix} and is therefore always non-negative; this implies that any extremum $y_*$ must satisfy $B_1(y_*)\ge 0$ and $B_3(y_*)\ge 0$.
+$A(y)$ is the ratio of the two Gaussians in Eq.~\ref{eq:LogNormalMix} and is therefore always non-negative; this implies that any extremum $y_*$ must satisfy $B_1(y_{*q})\ge 0$ and $B_3(y_{*p})\ge 0$.
 We note that the condition for extrema in $Q(I)$ requires that $\frac{d}{dy}P(y)=P(y)$ whereas, of course, extremizing $P(\log I)$ sets its derivative to zero, demonstrating the fact that the loci of the modes for $P(\log I)$ and $Q(I)$ are manifestly different.
-The region where the log-space distribution shows a second mode occurs when for $B_3(y^*)=A(y^*)$ admits three solutions whereas $B_1(y^*)=A(y^*)$ admits only one solution.
+The region where the log-space distribution shows a second mode occurs when for $B_3(y_{*p})=A(y_{*p})$ admits three solutions whereas $B_1(y_{*q})=A(y_{*q})$ admits only one solution.
 Given that these equations are transcendental, a graphical way to asses the number of solutions is to plot $A(y),B_1(y),B_3(y)$ and count the number of times $B_1$ and $B_3$ intersect $A$.\smallskip
@@
   \includegraphics[clip,width=1\hsize]{Fig2.png}
-  \caption{\textbf{Graphical solution to count the number of extrema.} When the red (blue) curves intersect the dashed black line, $P(\log I)$ ($Q(I)$) are extremized.
+  \caption{\textbf{Graphical solution to count the number of extrema.} We test our peak counting method, based on the same equations as in Fig.~\ref{fig:PlotsLinearLog}, but with additional settings.
+(Vertical axis represents the functions $A, B_1,B_3$ according to the legend).
+When the red (blue) curves intersect the dashed black line, $P(\log I)$ ($Q(I)$) are extremized.
 Dashed black: $A(y)$, blue: $B_1(y)$ and red: $B_3(y)$.
@@
 This leads to a simple criterion to determine whether $P(\log I)$ itself admits one or two modes - previously in Fig.~\ref{fig:PlotsLinearLog}(right) we saw an example where $P(\log I)$ is unimodal despite being generated from a mixture.
-Graphically, we see that for $B_3=F$ to have 3 solutions, $\log B_3(y)$ has to have a slope less than $\log A(y)$ about the extremum $y_*$.
-In other words, $\left.\frac{d}{dy} \log B_3(y)\right\vert_{y_*} \le \left.\frac{d}{dy} \log A(y)\right\vert_{y_*}$, with equality as the threshold between 1 and 3 extrema.
+Graphically, we see that for $B_3=A$ to have 3 solutions, $\log B_3(y)$ has to have a slope less than $\log A(y)$ about the extremum $y_{*p}$.
+In other words, $\left.\frac{d}{dy} \log B_3(y)\right\vert_{y_{*p}} \le \left.\frac{d}{dy} \log A(y)\right\vert_{y_{*p}}$, with equality as the threshold between 1 and 3 extrema.
 This leads to the following intuitive criterion,
@@
 \label{eq:Plogy3Maxima}
-(y_* - y_1)(y_2 - y_*) \ge \sigma ^2 \,\,\Longrightarrow \,\, \mbox{3 extrema for } P(\log I)\,,
+(y_{*p} - y_1)(y_2 - y_{*p}) \ge \sigma ^2 \,\,\Longrightarrow \,\, \mbox{3 extrema for } P(\log I)\,,
 \ee
-which states that for $P(\log I)$ to appear bimodal, it must have an extremum ($y_*$) such that the variance of the individual Gaussian components of $P(\log I)$ must be smaller than the distance between $y_*$ and the Gaussian centers.
-Substituting for $y_* \approx \log 316$, $y_1 = \log 100$ and $y_2=\log 1000$ and $\sigma^2 = 1.44$ we see that the criterion in Eq.~\ref{eq:Plogy3Maxima} is not satisfied and indeed in Fig.~\ref{fig:PlotsLinearLog}(right) and Fig.~\ref{fig:ExtremaSolutions}(bottom right) we see that $P(\log I)$ has only one mode.
-A similar condition can be derived for $Q(I)$, that is, $\left.\frac{d}{dy} \log B_1(y)\right\vert_{y_*} \le \left.\frac{d}{dy} \log A(y)\right\vert_{y_*}$, such that,
+which states that for $P(\log I)$ to appear bimodal, it must have an extremum ($y_{*p}$) such that the variance of the individual Gaussian components of $P(\log I)$ must be smaller than the distance between $y_{*p}$ and the Gaussian centers.
+Substituting for $y_{*p} \approx \log 316$, $y_1 = \log 100$ and $y_2=\log 1000$ and $\sigma^2 = 1.44$ we see that the criterion in Eq.~\ref{eq:Plogy3Maxima} is not satisfied and indeed in Fig.~\ref{fig:PlotsLinearLog}(right) and Fig.~\ref{fig:ExtremaSolutions}(bottom right) we see that $P(\log I)$ has only one mode.
+A similar condition can be derived for $Q(I)$, that is, $\left.\frac{d}{dy} \log B_1(y)\right\vert_{y_{*q}} \le \left.\frac{d}{dy} \log A(y)\right\vert_{y_{*q}}$, such that,
 \be 
 \label{eq:QI3Maxima}
-(y_* - y_1+\sigma^2)(y_2 - y_* - \sigma^2) \ge \sigma ^2\,\,\Longrightarrow \,\,\mbox{3 extrema for } Q(I)\,,
+(y_{*q} - y_1+\sigma^2)(y_2 - y_{*q} - \sigma^2) \ge \sigma ^2\,\,\Longrightarrow \,\,\mbox{3 extrema for } Q(I)\,,
 \ee \smallskip
-it is, however, hard to compare the two bounds analytically because the $y_*$ which extremizes $P(\log I)$ is different from the $y_*$ which extremizes $Q(I)$.
+it is, however, hard to compare the two bounds analytically because the $y_{*p}$ which extremizes $P(\log I)$ is different from the $y_{*q}$ which extremizes $Q(I)$.
 
-\newpage
+\subsection*{Experimental methods}
+All data and MATLAB scripts used for the experimental part of this work, together with accompanying MIFlowCyt checklists detailing the data analysis, are freely available online, as follows:
+
+\smallskip
+\begin{itemize}
+\item All MATLAB scripts used to generate the plots and analyze the data in {\small \href{https://github.com/AmirErez/BimodalLogspaceCytA/tree/master/Scripts}{github.com/AmirErez/BimodalLogspaceCytA/tree/master/Scripts}}
+\item Al data used in the manuscript and accompanying MIFlowCyt checklists come from two experiments:
+	\begin{enumerate}
+	\item 20140920-OT1-dynamics - used for Fig. 3 in the manuscript.
+	\item FeinermanScience2008 - previously published in Feinerman et al. Science Vol. 321, Issue 5892, pp. 1081-1084 (2008). 
+	\end{enumerate}
+These data with their respective MIFlowCyt checklists are available in {\small
+\href{https://github.com/AmirErez/BimodalLogspaceCytA/tree/master/Data}{github.com/AmirErez/BimodalLogspaceCytA/tree/master/Data}}
+\end{itemize} 
+All are accessible in {\small\href{https://github.com/AmirErez/BimodalLogspaceCytA}{github.com/AmirErez/BimodalLogspaceCytA}}.
+We briefly sketch the methods for each of the two experiments:
+\subsubsection*{Experiment 1:} Primary mouse T lymphocytes (OT-1) were activated ex vivo and cultured.
+OT-1 cells were activated by peptide (SIINFEKL) treated and irradiated (3,000 RAD) primary antigen presenting cells (APCs) from a C57BL/6 mouse.
+OT-1 cells were labeled with an amine-reactive dye, CTV, according to the manufacture’s protocol (Molecular Probes) for in silico identification.
+Cells were allowed to rest for one hour after CTV staining, and then distributed in a 96-well v-bottom plate.
+Cells were then incubated with unique doses of either SRC or MEK inhibitor at a temperature of $37^o$C for 5 minutes.
+After which,  we added the peptide pulsed APCs (10 APCs to 1 OT-1 T cell) and pelleted mixture of cells by centrifugation for 10 s at 460 rcf at room temperature.
+The pellet was then incubated at a temperature of $37^o$C for 10 minutes, followed by 15 minutes of chemical fixing with 2\% PFA on ice and then permeabilization with ice cold 90\% MeOH.
+Samples were kept at a temperature of $-20^o$C until labeling for FCM.
+
+\subsubsection*{Experiment 2:} C57BL/6N splenocytes were pulsed for 2hr with 100nM SIINFEKL peptide, then irradiated (3000RAD), washed once and used as stimulator/feeder cells.
+OT-1 lymphocytes were harvested from auxiliary, brachial and inguinal lymph nodes as well as spleen (splenocytes were treated with ACK lysis buffer to remove red blood cells), and mixed with SIINFEKL-pulsed C57BL/6N splenocytes in complete RPMI.
+After two days, cells were expanded by diluting 2 fold into medium containing 100 pM IL-2.
+After four days, the cells were again expanded by 2 fold dilution into medium with IL-2.
+After one more day of culture, cells were harvested and spun through a 1.09 density Ficoll- Paque Plus gradient (GE Healthcare) to remove dead cells.
+Live cells were recovered, washed twice in complete medium and resuspended at 1 million/ml in complete medium with 100pM IL-2.
+Cells were used for experiments between 6 and 8 days after primary stimulation.
+The ppERK response to SRC and MEK inhibition was measured using primary OT-1 T-lymphocytes activated with RMA-S APCs.
+RMA-S cells were suspended in culture with 1 nM SIINFEKL peptide for 2 h at 37 $^o$C, 5\% CO\textsubscript{2}, and on a rotator to guarantee mixing.
+During this time we labelled OT-1 cells with an amine-reactive dye, CTV, according to the manufacture’s protocol (Molecular Probes).
+We rested the OT-1 cells one hour after CTV staining, and then distributed them in a 96-well v-bottom plate.
+Each well was given various doses of SRC inhibitor and MEK inhibitor and kept at 37 $^o$C for 5 min.
+Following the 5 min exposure to the inhibitors, we added the peptide pulsed RMA-S (10 RMA-S to 1 OT-1 T cell) and pelleted by centrifugation for 10 s at 460 rcf at room temperature.
+This step guaranteed that both cell types, OT-1 and RMA-S, came into contact.
+The cells were allowed to activate for 10 min, followed by fixing on ice in 2\% PFA, and then permeabilized and stored in 90\% MeOH at  -20 $^o$C.
+APCs were pulsed with serial dilutions of OVA or variant peptides for 2 hr at 37$^o$C, then washed with T cell medium at the time of harvest, and resuspended with anti-CD8α (53-6.72)-Fab-coated T cells in their conditioned media in a V-bottom 96-well plate (Corning).
+Fab-coating was performed 10 min before cell use by incubating T cells with 10 μg/ml of Fab fragment.
+T:APC cell contacts were synchronized using a quick centrifugal spin (10s at 400g).
+Plates containing T:APC conjugates were placed on a water bath at 37$^o$C and incubated for 5 min.
+Supernatants were then discarded, T:APC conjugates disrupted by vortexing, and cells resuspended in ice-cold 4\% paraformaldehyde for 15 min.
+Cells were then permeabilized with ice-cold 90\% methanol for 15 min on ice, and washed twice with FACS buffer.
 
@@
 
-We now apply our analysis to experimental data, namely the measured distribution of Extracellular Signal-regulated Kinase (ERK) phosphorylation (ppERK) signaling in CD8+ primary mouse T-cells responding to antigens and inhibited by the SRC inhibitor Dasatinib.
-These data were acquired in exactly the same manner as the experiments in Ref.~\cite{Vogel2016}, for brevity we refer the reader there for all experimental details.
+We now apply our analysis to experimental data, by analyzing two different experiments, described briefly in the \emph{Methods} section and in more detail in the MIFlowCyt checklists available online.
+Both experiments measure the distribution of Extracellular Signal-regulated Kinase (ERK) phosphorylation (ppERK) signaling in CD8+ primary mouse T-cells responding to antigens.
+
+\smallskip
+In experiment 1, the T-cells were inhibited by the SRC inhibitor Dasatinib.
 In Fig.~\ref{fig:Experiment}(A) we see how a commercially-available analysis software (FlowJo \cite{FlowJo}) plots the distribution of ppERK in such an experiment, which clearly shows a bimodal structure.
@@
 Hartigan's unimodality p-values for $\log I$ (red) and $I$ (blue) are taken directly from the data without binning, corroborating that $\log I$ is bimodal whereas $I$ is unimodal.
-\textbf{(C)} Graphic solution of the extrema conditions as in Fig.~\ref{fig:ExtremaSolutions} explicitly reveals the three solutions for $P(\log I_*)$ (red line intersects black dashed) as opposed to the single solution for $Q(I_*)$ (the blue line intersects the black dashed line up beyond the plotted area, solution also plotted as blue star in the middle plot), indicating that $Q(I)$ has only one mode.}
+\textbf{(C)} (Vertical axis represents the functions $A, B_1,B_3$ according to the legend).
+Graphic solution of the extrema conditions as in Fig.~\ref{fig:ExtremaSolutions} explicitly reveals the three solutions for $P(\log I_*)$ (red line intersects black dashed) as opposed to the single solution for $Q(I_*)$ (the blue line intersects the black dashed line up beyond the plotted area, solution also plotted as blue star in the middle plot), indicating that $Q(I)$ has only one mode.}
   \label{fig:Experiment}
@@
 \smallskip
+We proceed to analyze experiment 2, these data were taken from Ref.~\cite{Feinerman2008}.
 We briefly sketch what follows: by considering both ERK1 and ppERK levels, (the two being biologically related and experimentally correlated), we demonstrate that dividing one by the other ameliorates the mismatch between the logarithmic and linear representations.
@@
 In Fig.~\ref{fig:ppERKvsERK1heatmap}(A), we show the experimental data we will use in our proposed solution.
-Here we returned to our single-cell measurements of ERK phosphorylation in primary mouse T cells in Ref.~\cite{Feinerman2008}.
 We show a heat map of the \emph{joint} distribution of ppERK ($I_{ppERK}$) and total ERK1 ($I_{ERK1}$) expression in mouse CD8+ T-cells.
@@
 The weak dependence between activation state and ERK1 levels is reasonable, if we account for cell-to-cell variability, since for a given stimulus some cells inevitably respond differently from the typical cell \cite{Spencer2009}.
+\smallskip
+
 We summarize the causal structure for this system in Fig.~\ref{fig:ppERKvsERK1heatmap}(B) which depicts a probabilistic graphical model \cite{Koller2009} of the flow of influence from cell-to-cell variability and activation signal, to ERK1 levels and activation status, and finally to the distribution of ppERK.
 We depict the pair ERK1-ppERK in a template (rectangles), to suggest to the reader the existence of multiple other pairs.
-Importantly, in what follows we show how to better resolve the log-space peak; this recipe, together with the model in Fig.~\ref{fig:ppERKvsERK1heatmap}(B) can be used \emph{a priori} in automatic gating and clustering algorithms to prevent some of the mismatch between logarithmic and linear binning strategies.
+Thus, by formalizing the flow of influence of cell-to-cell variability on the observed fluorescence signal we provide reasoning for what follows.
+\smallskip
+
+We proceed to show how to better resolve the log-space peak; this recipe, together with the model in Fig.~\ref{fig:ppERKvsERK1heatmap}(B) can be used \emph{a priori} in automatic gating and clustering algorithms to prevent some of the mismatch between logarithmic and linear binning strategies.
 For stochastic modeling, such a structure presents an opportunity to analyze the structure and propagation of noise in the system \cite{Prill2015,Ching2017}.
@@
 \textbf{(C)} The same treatment but for $\tilde{I} = I_{ppERK}/I_{ERK1}$, \textbf{(D)} shows that both $\tilde{I}$ and $\log \tilde{I}$ have two modes, thus normalizing ppERK levels by total ERK1 maintains the bimodal structure both in $P(\log \tilde{I})$ and in $Q(\tilde{I})$.
-}
+In (B) and (D), the vertical axis represents the functions $A, B_1,B_3$ according to the legend.}
   \label{fig:NormppERKbyERK}
@@
 This is a simple way to introduce knowledge of the biological structure of the data into otherwise objective clustering algorithms, without compromising their objectivity.
-We propose and test some features of a graphical model that captures the structure of such dependencies in a way potentially useful for those interested in automatic gating and clustering algorithms.
+We propose and test some features of a graphical model that captures the structure of such dependencies and the propagation of noise from cell-to-cell variability into the observed fluorescence signal.
+The graphical model is used to motivate our suggestion of how to ameliorate the mismatch between the linear and logarithmic representations.
+Moreover, it is potentially useful for those interested in correctly capturing these dependencies for automatic gating and clustering algorithms.
 

### 2017-11-17 Manuscript/LogSpaceDistributionsCytometryA.tex
Minor edits

@@
 
-\documentclass[11pt,a4paper,final]{article}
+\documentclass[11pt,a4paper,draft]{article}
 \usepackage{setspace}
@@
 
-\section*{Methods - theory}
-We focus this \emph{Methods} section on the theoretical aspect of this work. Brief details regarding the experimental methods can be found at the end of this section.
+\section*{Methods}
+In this \emph{Methods} section we present, in two detailed sections, the theoretical and experiment techniques used in our study.
+
+\subsection*{Theory}
 
@@
  \centering
-  \includegraphics[clip,width=1\hsize]{Fig1.png}
+%  \includegraphics[clip,width=1\hsize]{Fig1.tif}
   \caption{\textbf{Log-normal mixture showing the mismatch in the number of peaks.} \textbf{Top row, red:} $P(\log I)$ vs.
@@
  \centering
-  \includegraphics[clip,width=1\hsize]{Fig2.png}
+%  \includegraphics[clip,width=1\hsize]{Fig2.tif}
   \caption{\textbf{Graphical solution to count the number of extrema.} We test our peak counting method, based on the same equations as in Fig.~\ref{fig:PlotsLinearLog}, but with additional settings.
@@
 Cells were allowed to rest for one hour after CTV staining, and then distributed in a 96-well v-bottom plate.
-Cells were then incubated with unique doses of either SRC or MEK inhibitor at a temperature of $37^o$C for 5 minutes.
+Cells were then incubated with unique doses of SRC inhibitor Dasatinib at a temperature of $37^o$C for 5 minutes.
 After which,  we added the peptide pulsed APCs (10 APCs to 1 OT-1 T cell) and pelleted mixture of cells by centrifugation for 10 s at 460 rcf at room temperature.
@@
  \centering
-  \includegraphics[width=1\linewidth]{Fig3.png}
+%  \includegraphics[width=1\linewidth]{Fig3.tif}
   \caption{\textbf{Analysis of experimental data reveals the effect we describe in a real scenario.
@@
     \centering
-    \includegraphics[width=1\linewidth]{Fig4.png}
+%    \includegraphics[width=1\linewidth]{Fig4.tif}
    \caption{\textbf{Analysis of experimental data with two correlated measurements.} \textbf{(A)} The joint distribution $P_2(\log I_{ppERK},\log I_{ERK1})$ as a heat map with its marginals plotted on the top and on its right.
@@
  \centering
-   \includegraphics[width=1\linewidth]{Fig5.png}  
+%   \includegraphics[width=1\linewidth]{Fig5.tif}  
   \caption{\textbf{By dividing ppERK readings by ERK1, we can ameliorate the mismatch between the two representations.} \textbf{(A)} $P(\log I_{ppERK})$ and $Q(I_{ppERK})$ vs.
@@
  \centering
- \includegraphics[clip,width=1\hsize]{Fig6.png}
+% \includegraphics[clip,width=1\hsize]{Fig6.tif}
   \caption{\textbf{Testing Hartigan's p-value $p_u$ for unimodality on a weakly bimodal log-normal mixture.} The heat-map shows $p_u$ as a function of the number of bootstrap tests and number of events in each test, for $\sigma_1=0.4$ and $\sigma_2=0.8$ as in Fig.~\ref{fig:ExtremaSolutions}(top,middle).
@@
  \centering
-   \includegraphics[width=1\linewidth]{Fig7.png}  
+%   \includegraphics[width=1\linewidth]{Fig7.tif}  
   \caption{{\bf Test for weak dependence of ERK1 and activation status.} Whereas independence implies that $P(\log ERK1) = P(\log ERK1 \vert Activation)$, in fact there is a weak dependence between them regardless of whether we employ a vertical (red, defined by threshold value $ppERK_*$) or a diagonal (grey) gate (the diagonal gate showing a weaker dependence); this is observed directly by noting that the different distributions in Fig.~\ref{S1_Fig} do not lie on top of each other.

### 2017-11-17 Manuscript/LogSpaceDistributionsCytometryA.tex
Submission

@@
 \usepackage{setspace}
-\usepackage{hyperref}
+%\usepackage{hyperref}
+%\usepackage[T1]{fontenc}
 \doublespacing
@@
 \begin{document}
-\def\be{\begin{equation}}
-\def\ee{\end{equation}}
-\def\bea{\begin{eqnarray}}
-\def\eea{\end{eqnarray}}
+%\def\be{\begin{equation}}
+%\def\ee{\end{equation}}
+%\def\bea{\begin{eqnarray}}
+%\def\eea{\end{eqnarray}}
 \def\nl{\\ \noindent}
@@
 Given the pdf $P(\log I)$, one can formally derive $Q(I)$ as \cite{Mukhopadhyay2000},
-\bea 
+\begin{eqnarray} 
 \label{eq:P2Q}
@@
  &=& \frac{1}{I}P(\log I) = e^{-y} P(y)\,,
-\eea
+\end{eqnarray}
 with $\log I\equiv y$.
@@
 We now address the question: when is $Q(I)$ unimodal while $P({\log I})$ is multimodal ?
+
 \smallskip
-
-Cytometry data is often amenable to modeling as a log-normal mixture ({\it e.g.} \cite{Vogel2016}).
+Cytometry data is often amenable to modeling as a log-normal mixture (e.g., \cite{Vogel2016}).
 To demonstrate the log/linear mismatch we consider a mixture of two populations, characterized by the distribution of intensity.
 We define $P(\log I)$ as follows:
-\be
+\begin{equation}
+P(\log I) = \frac{(1-\alpha)}{\sqrt{2\pi\sigma_1^2}}e^{-\frac{(\log I - y_1)^2}{2\sigma_1^2}} + \frac{\alpha}{\sqrt{2\pi\sigma_2^2}}e^{-\frac{(\log I - y_2)^2}{2\sigma_2^2}},
 \label{eq:LogNormalMix}
-P(\log I) = \frac{(1-\alpha)}{\sqrt{2\pi\sigma_1^2}}e^{-\frac{(\log I - y_1)^2}{2\sigma_1^2}} + \frac{\alpha}{\sqrt{2\pi\sigma_2^2}}e^{-\frac{(\log I - y_2)^2}{2\sigma_2^2}},
-\ee
+\end{equation}
 with $y_{1,2}=\log I_{1,2}$ which are the loci of the centers of the left and right Gaussians in log-space, respectively, and $\sigma_{1,2}$ the log-space standard deviations.
@@
 This is achieved by recasting Eq.~\ref{eq:LogNormalMix} and its derivatives into the functions $A(y),B_1(y),B_3(y)$, defined as,
-\bea 
+\begin{eqnarray} 
 \label{eq:extrema}
@@
 B_3(y) &=& \frac{(y-y_1)/\sigma_1^2}{(y_2-y)/\sigma_2^2}\,.
-\eea
+\end{eqnarray}
 The outcome of this approach is that, to find the extrema of $Q(I)$, one must find the solutions $y_{*q}$ for $B_1(y_{*q})=A(y_{*q})$.
@@
 This leads to the following intuitive criterion,
-\be 
+\begin{equation} 
+(y_{*p} - y_1)(y_2 - y_{*p}) \ge \sigma ^2 \,\,\Longrightarrow \,\, \mbox{3 extrema for } P(\log I)\,,
 \label{eq:Plogy3Maxima}
-(y_{*p} - y_1)(y_2 - y_{*p}) \ge \sigma ^2 \,\,\Longrightarrow \,\, \mbox{3 extrema for } P(\log I)\,,
-\ee
+\end{equation}
 which states that for $P(\log I)$ to appear bimodal, it must have an extremum ($y_{*p}$) such that the variance of the individual Gaussian components of $P(\log I)$ must be smaller than the distance between $y_{*p}$ and the Gaussian centers.
@@
 A similar condition can be derived for $Q(I)$, that is, $\left.\frac{d}{dy} \log B_1(y)\right\vert_{y_{*q}} \le \left.\frac{d}{dy} \log A(y)\right\vert_{y_{*q}}$, such that,
-\be 
+\begin{equation} 
+(y_{*q} - y_1+\sigma^2)(y_2 - y_{*q} - \sigma^2) \ge \sigma ^2\,\,\Longrightarrow \,\,\mbox{3 extrema for } Q(I)\,,
 \label{eq:QI3Maxima}
-(y_{*q} - y_1+\sigma^2)(y_2 - y_{*q} - \sigma^2) \ge \sigma ^2\,\,\Longrightarrow \,\,\mbox{3 extrema for } Q(I)\,,
-\ee \smallskip
+\end{equation}
+
+\smallskip
 it is, however, hard to compare the two bounds analytically because the $y_{*p}$ which extremizes $P(\log I)$ is different from the $y_{*q}$ which extremizes $Q(I)$.
@@
 \begin{itemize}
-\item All MATLAB scripts used to generate the plots and analyze the data in {\small \href{https://github.com/AmirErez/BimodalLogspaceCytA/tree/master/Scripts}{github.com/AmirErez/BimodalLogspaceCytA/tree/master/Scripts}}
+\item All MATLAB scripts used to generate the plots and analyze the data in 
+github.com/AmirErez/BimodalLogspaceCytA/tree/master/Scripts 
 \item Al data used in the manuscript and accompanying MIFlowCyt checklists come from two experiments:
@@
 	\end{enumerate}
-These data with their respective MIFlowCyt checklists are available in {\small
-\href{https://github.com/AmirErez/BimodalLogspaceCytA/tree/master/Data}{github.com/AmirErez/BimodalLogspaceCytA/tree/master/Data}}
+These data with their respective MIFlowCyt checklists are available in 
+github.com/AmirErez/BimodalLogspaceCytA/tree/master/Data 
 \end{itemize} 
-All are accessible in {\small\href{https://github.com/AmirErez/BimodalLogspaceCytA}{github.com/AmirErez/BimodalLogspaceCytA}}.
+All are accessible in 
+https://github.com/AmirErez/BimodalLogspaceCytA
 We briefly sketch the methods for each of the two experiments:
@@
 Cells were allowed to rest for one hour after CTV staining, and then distributed in a 96-well v-bottom plate.
-Cells were then incubated with unique doses of SRC inhibitor Dasatinib at a temperature of $37^o$C for 5 minutes.
+Cells were then incubated with unique doses of SRC inhibitor Dasatinib at a temperature of 37C for 5 minutes.
 After which,  we added the peptide pulsed APCs (10 APCs to 1 OT-1 T cell) and pelleted mixture of cells by centrifugation for 10 s at 460 rcf at room temperature.
-The pellet was then incubated at a temperature of $37^o$C for 10 minutes, followed by 15 minutes of chemical fixing with 2\% PFA on ice and then permeabilization with ice cold 90\% MeOH.
-Samples were kept at a temperature of $-20^o$C until labeling for FCM.
+The pellet was then incubated at a temperature of 37C for 10 minutes, followed by 15 minutes of chemical fixing with 2\% PFA on ice and then permeabilization with ice cold 90\% MeOH.
+Samples were kept at a temperature of -20C until labeling for FCM.
 
@@
 The ppERK response to SRC and MEK inhibition was measured using primary OT-1 T-lymphocytes activated with RMA-S APCs.
-RMA-S cells were suspended in culture with 1 nM SIINFEKL peptide for 2 h at 37 $^o$C, 5\% CO\textsubscript{2}, and on a rotator to guarantee mixing.
+RMA-S cells were suspended in culture with 1 nM SIINFEKL peptide for 2 h at 37C, 5\% CO\textsubscript{2}, and on a rotator to guarantee mixing.
 During this time we labelled OT-1 cells with an amine-reactive dye, CTV, according to the manufacture’s protocol (Molecular Probes).
 We rested the OT-1 cells one hour after CTV staining, and then distributed them in a 96-well v-bottom plate.
-Each well was given various doses of SRC inhibitor and MEK inhibitor and kept at 37 $^o$C for 5 min.
+Each well was given various doses of SRC inhibitor and MEK inhibitor and kept at 37C for 5 min.
 Following the 5 min exposure to the inhibitors, we added the peptide pulsed RMA-S (10 RMA-S to 1 OT-1 T cell) and pelleted by centrifugation for 10 s at 460 rcf at room temperature.
 This step guaranteed that both cell types, OT-1 and RMA-S, came into contact.
-The cells were allowed to activate for 10 min, followed by fixing on ice in 2\% PFA, and then permeabilized and stored in 90\% MeOH at  -20 $^o$C.
-APCs were pulsed with serial dilutions of OVA or variant peptides for 2 hr at 37$^o$C, then washed with T cell medium at the time of harvest, and resuspended with anti-CD8α (53-6.72)-Fab-coated T cells in their conditioned media in a V-bottom 96-well plate (Corning).
+The cells were allowed to activate for 10 min, followed by fixing on ice in 2\% PFA, and then permeabilized and stored in 90\% MeOH at  -20C.
+APCs were pulsed with serial dilutions of OVA or variant peptides for 2 hr at 37C, then washed with T cell medium at the time of harvest, and resuspended with anti-CD8α (53-6.72)-Fab-coated T cells in their conditioned media in a V-bottom 96-well plate (Corning).
 Fab-coating was performed 10 min before cell use by incubating T cells with 10 μg/ml of Fab fragment.
 T:APC cell contacts were synchronized using a quick centrifugal spin (10s at 400g).
-Plates containing T:APC conjugates were placed on a water bath at 37$^o$C and incubated for 5 min.
+Plates containing T:APC conjugates were placed on a water bath at 37C and incubated for 5 min.
 Supernatants were then discarded, T:APC conjugates disrupted by vortexing, and cells resuspended in ice-cold 4\% paraformaldehyde for 15 min.
@@
 %
-\bea \nonumber
+\begin{eqnarray} \nonumber
   \left.\frac{d P(y)}{dy} \right\vert_{y=y_*}&=&
@@
   &=& 0\nonumber
-\eea
+\end{eqnarray}
 %
@@
 %
-\be 
-  \label{eq:extremaP}
+\begin{equation} 
   \underbrace{\frac{(y_* - y_1) / \sigma_1^2}{(y_2-y_*)/\sigma_2^2}}_{B_3(y_*)} =
@@
     e^{\frac{(y_*-y1)^2}{2\sigma_1 ^2}}\gauss{2}{y_2}{y_*}}_{A(y_*)}\,,
-\ee
+\label{eq:extremaP}
+\end{equation}
 %
@@
 %
-\be
+\begin{equation}
+P(y_*) = \left.\frac{d P(y)}{dy}\right\vert_{y = y_*}\,,
 \label{eq:QExtremaStrange}
-P(y_*) = \left.\frac{d P(y)}{dy}\right\vert_{y = y_*}\,,
-\ee
+\end{equation}
 %
@@
 %
-\be
+\begin{equation}
+  \underbrace{\frac{\frac{y_*-y_1}{\sigma_1^2}+1}{\frac{y_2-y_*}{\sigma_2^2}-1}}_{B_1(y_*)} = A(y_*) 
   \label{eq:extremaQ}
-  \underbrace{\frac{\frac{y_*-y_1}{\sigma_1^2}+1}{\frac{y_2-y_*}{\sigma_2^2}-1}}_{B_1(y_*)} = A(y_*) 
-\ee
+\end{equation}
 %

### 2018-01-09 Manuscript/LogSpaceDistributionsCytometryA.tex
Small changes

@@
   \caption{\textbf{Log-normal mixture showing the mismatch in the number of peaks.} \textbf{Top row, red:} $P(\log I)$ vs.
-$\log I$ normalized to max.
+$\log I$ normalized to maximum probability.
 \textbf{Middle row, magenta:} $Q(I)$ vs.
-$I$ normalized to max, and plotted on a narrower range.
+$I$ normalized to maximum probability, and plotted on a narrower range.
 \textbf{Bottom row, blue:} $Q(I)$ vs.
-$\log I$ normalized to max, rescaling the x-axis as in Ref.~\cite{Novo2008}.
+$\log I$ normalized to maximum probability, rescaling the x-axis as in Ref.~\cite{Novo2008}.
 In the central column where $\sigma_1=\sigma_2=0.8$, $P(\log I)$ shows explicit bimodality whereas $Q(I)$ is unimodal.
@@
 $A(y)$ is the ratio of the two Gaussians in Eq.~\ref{eq:LogNormalMix} and is therefore always non-negative; this implies that any extremum $y_*$ must satisfy $B_1(y_{*q})\ge 0$ and $B_3(y_{*p})\ge 0$.
-We note that the condition for extrema in $Q(I)$ requires that $\frac{d}{dy}P(y)=P(y)$ whereas, of course, extremizing $P(\log I)$ sets its derivative to zero, demonstrating the fact that the loci of the modes for $P(\log I)$ and $Q(I)$ are manifestly different.
+In the \emph{Supporting information}, we derive that the condition for extrema in $Q(I)$ requires that $\frac{d}{dy}P(y)=P(y)$ whereas, of course, extremizing $P(\log I)$ sets its derivative to zero, demonstrating the fact that the loci of the modes for $P(\log I)$ and $Q(I)$ are manifestly different.
 The region where the log-space distribution shows a second mode occurs when for $B_3(y_{*p})=A(y_{*p})$ admits three solutions whereas $B_1(y_{*q})=A(y_{*q})$ admits only one solution.
@@
 github.com/AmirErez/BimodalLogspaceCytA/tree/master/Scripts 
-\item Al data used in the manuscript and accompanying MIFlowCyt checklists come from two experiments:
+\item All data used in the manuscript and accompanying MIFlowCyt checklists come from two experiments:
 	\begin{enumerate}
@@
 The cells were allowed to activate for 10 min, followed by fixing on ice in 2\% PFA, and then permeabilized and stored in 90\% MeOH at  -20C.
-APCs were pulsed with serial dilutions of OVA or variant peptides for 2 hr at 37C, then washed with T cell medium at the time of harvest, and resuspended with anti-CD8α (53-6.72)-Fab-coated T cells in their conditioned media in a V-bottom 96-well plate (Corning).
-Fab-coating was performed 10 min before cell use by incubating T cells with 10 μg/ml of Fab fragment.
+APCs were pulsed with serial dilutions of OVA or variant peptides for 2 hr at 37C, then washed with T cell medium at the time of harvest, and resuspended with anti-CD8 (53-6.72)-Fab-coated T cells in their conditioned media in a V-bottom 96-well plate (Corning).
+Fab-coating was performed 10 min before cell use by incubating T cells with 10 $\mu$g/ml of Fab fragment.
 T:APC cell contacts were synchronized using a quick centrifugal spin (10s at 400g).
@@
 Our procedure for estimating $P(\log I)$ gives the two modes as in FlowJo (red dots) whereas $Q(I)$ (derived using Eq.~\ref{eq:P2Q}) has a single mode (blue dots).
+It also agrees with the results of Hartigan's unimodality p-values, $p_U$, explained in more detail in the \emph{Supporting information}.
 We fit $P(\log I)$ as a Gaussian mixture.
@@
   \label{fig:NormppERKbyERK}
+\end{figure}
+
+\smallskip
+
+\begin{figure}[!htb]
+ \centering
+%   \includegraphics[width=1\linewidth]{Fig7.tif}  
+  \caption{{\bf Test for weak dependence of ERK1 and activation status.} Whereas independence implies that $P(\log ERK1) = P(\log ERK1 \vert Activation)$, in fact there is a weak dependence between them regardless of whether we employ a vertical (red, defined by threshold value $ppERK_*$) or a diagonal (grey) gate (the diagonal gate showing a weaker dependence); this is observed directly by noting that the different distributions in Fig.~\ref{S1_Fig} do not lie on top of each other.
+To quantify this difference, the inset shows the mutual information between $P(\log ERK1)$ and $P(\log ERK1 | Activation)$, with the circle pointing out the particular concentration of stimulus (out of all concentrations used in Ref.~\cite{Feinerman2008}) we chose to plot in this example.
+The chosen concentration has the highest mutual information, {\it i.e.}, the lowest ability to discern between the active and inactive states.}
+  \label{S1_Fig}
 \end{figure}
@@
 
-\subsection*{Relation between ERK1 level and activation status}
-\begin{figure}[!htb]
- \centering
-%   \includegraphics[width=1\linewidth]{Fig7.tif}  
-  \caption{{\bf Test for weak dependence of ERK1 and activation status.} Whereas independence implies that $P(\log ERK1) = P(\log ERK1 \vert Activation)$, in fact there is a weak dependence between them regardless of whether we employ a vertical (red, defined by threshold value $ppERK_*$) or a diagonal (grey) gate (the diagonal gate showing a weaker dependence); this is observed directly by noting that the different distributions in Fig.~\ref{S1_Fig} do not lie on top of each other.
-To quantify this difference, the inset shows the mutual information between $P(\log ERK1)$ and $P(\log ERK1 | Activation)$, with the circle pointing out the particular concentration of stimulus (out of all concentrations used in Ref.~\cite{Feinerman2008}) we chose to plot in this example.
-The chosen concentration has the highest mutual information, {\it i.e.}, the lowest ability to discern between the active and inactive states.}
-  %\caption{\textbf{By dividing ppERK readings by ERK1, we can ameliorate the mismatch between the two representations.} \textbf{(A)} $P(\log I)$ and $Q(I)$ vs.
-$\log I$ (dots: data, lines: gaussian mixture fit) together with \textbf{(B)} their extrema analysis, showing that the second mode in log ppERK does not exist if the data is linearly binned.
-\textbf{(C)} The same treatment but for $\tilde{I} = I_{ppERK}/I_{ERK1}$, \textbf{(D)} shows that both $\tilde{I}$ and $\log \tilde{I}$ have two modes, thus normalizing ppERK levels by total ERK1 maintains the bimodal structure both in $P(\log \tilde{I})$ and in $Q(\tilde{I})$.
-}
-  \label{S1_Fig}
-\end{figure}
+%\subsection*{Relation between ERK1 level and activation status}
 
